---
title: "Categorical Responses"
description: "Try the GLMcat package and drmOrdinal"
author: "Zhenglei Gao"
date: "01/10/2025"
categories:
  - code
  - analysis
editor_options: 
  chunk_output_type: console
format:
  html:
    code-fold: true
---

GLMcat is an R package that encompasses lots of models specified in a similar way: (ratio, cdf, design: parallel or complete).

ratio can be cumulative, sequential, or adjacent for ordinal data, reference for nominal data. Logistic, normal, Cauchy, Student, Gompertz and Gumbel are the commonly used link functions or CDF. Note that Gumbel cdf is the symmetric of the Gompertz cdf. 

For parallel design, the effects of the covariates are assumed to be constant across the response
categories, whereas for complete design, the covariates effects are different across different response categories. The linear predictios are $\eta_j=\alpha_j+x'\delta_j$ and $\eta_j=\alpha_j+x'\delta$ for $j$ in $(1,...,J-1)$

There are several type of models for categorical data:

- multinomial  (reference, logistic, complete) --> devoted to nominal instead of ordinal
- odds parallel logit (cumulative, logistic, parallel)
- parallel hazard (sequential, gompertz, parallel)
- continuation ratio logit model (sequential, logistic, complete)
- adjacent logit model (adjacent, logistic, complete)

```{r include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
})
```


## The dataset used in a previous post

The data was generated by Changjian using SAS Probnorm with Vx=0.25 and Vr=0.15, a=-2.5, b=1.0, ER50=12.18. 
Model: probit(y) = a + b*logdose + ri;

```{r}
pnorm(-2.5+1*log(c(2L, 4L, 8L,16L, 32L,  64L)))

```


```{r collapse=TRUE}
dattab_new <- read.table(textConnection("Obs	rep	dose	yt  y0 
1	1	2	0.02021	0
2	2	2	0.02491	0
3	3	2	0.00760	0
4	5	2	0.04466	0
5	6	2	0.00037	0
6	8	2	0.05386	0
7	9	2	0.07205	0
8	10	2	0.01125	0
9	1	4	0.02011	0
10	7	4	0.09058	0
11	8	4	0.06255	0
12	10	8	0.09431	0
13	4	2	0.14060	A
14	7	2	0.22223	A
15	2	4	0.20943	A
16	3	4	0.20959	A
17	4	4	0.17305	A
18	9	4	0.11405	A
19	10	4	0.17668	A
20	1	8	0.12756	A
21	6	8	0.11478	A
22	6	32	0.20602	A
23	5	4	0.26650	B
24	6	4	0.27344	B
25	3	8	0.27021	B
26	5	8	0.30662	B
27	8	8	0.29319	B
28	9	8	0.37300	B
29	6	16	0.36224	B
30	9	16	0.31316	B
31	2	8	0.55845	C
32	4	8	0.44811	C
33	3	16	0.42677	C
34	3	32	0.52315	C
35	7	8	0.67080	D
36	2	16	0.71776	D
37	4	16	0.73038	D
38	5	16	0.64232	D
39	7	16	0.68720	D
40	8	16	0.61088	D
41	5	32	0.72342	D
42	8	32	0.63594	D
43	1	16	0.77171	E
44	10	16	0.74087	E
45	2	32	0.79477	E
46	4	32	0.88546	E
47	7	32	0.78002	E
48	9	32	0.81456	E
49	10	32	0.89465	E
50	1	32	0.96129	F
51	1	64	0.96127	F
52	2	64	0.91687	F
53	3	64	0.97204	F
54	4	64	0.99268	F
55	5	64	0.98935	F
56	6	64	0.96263	F
57	7	64	0.95435	F
58	8	64	0.92081	F
59	9	64	0.91776	F
60	10	64	0.99104	F"),header = TRUE)
```


```{r collapse=FALSE}
dattab_new <- dattab_new %>% mutate(yy = as.numeric(plyr::mapvalues(y0,from = c("0","A","B","C","D","E","F"),to = c(0,10,30,50,70,90,100)))/100) %>% mutate(yy2 = as.numeric(plyr::mapvalues(y0,from = c("0","A","B","C","D","E","F"),to = c(0.05,0.18,0.34,0.50,0.66,0.82,0.95))))

ftable(xtabs(~ y0 + dose, data = dattab_new)) ##%>% gt::gt() ## as.data.frame(.) %>% knitr::kable(.,digits = 3)
```


```{r}


dattab_new %>% group_by(y0) %>% summarise(n=n(),meany=mean(yt),meanyy2=mean(yy2)) %>% knitr::kable(.,digits = 3)

dattab_new %>% group_by(dose) %>% summarise(n=n(),meany=mean(yt),meanyy=mean(yy)) %>% knitr::kable(.,digits = 3)
```


```{r}
ggplot(dattab_new,aes(x = dose, y=yt))+geom_point() +  geom_smooth(method="glm", method.args=list(family="quasibinomial"), formula="y ~ log(x)",
                       se =TRUE, size=1.5)
```


| y0   | dose |  2 |  4 |  8 | 16 | 32 | 64 |
|:-----|------|----:|----:|----:|----:|----:|----:|
| 0    |      |   8 |   3 |   1 |   0 |   0 |   0 |
| A    |      |   2 |   5 |   2 |   0 |   1 |   0 |
| B    |      |   0 |   2 |   4 |   2 |   0 |   0 |
| C    |      |   0 |   0 |   2 |   1 |   1 |   0 |
| D    |      |   0 |   0 |   1 |   5 |   2 |   0 |
| E    |      |   0 |   0 |   0 |   2 |   5 |   0 |
| F    |      |   0 |   0 |   0 |   0 |   1 |  10 |



```{r}
## fit ordered logit model and store results 'm'
dattab_new $y0 <- factor(dattab_new$y0, levels = c("0","A","B","C","D","E","F"),ordered = TRUE)

m <- MASS::polr(y0 ~ log(dose), data = dattab_new, Hess=TRUE)
summary(m)
ctable <- coef(summary(m))

## At ER50, the cumulative probability probability of the response being in a higher category is close to 1.
plogis(ctable[,1] + ctable[,2]*log(12.18))
## calculate and store p values
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2

## combined table
(ctable <- cbind(ctable, "p value" = p))

(ci <- confint(m)) ## profiled CI

exp(cbind(coef(m),t(ci)))
## OR and CI
exp(cbind(OR = coef(m), ci))

newdat <- data.frame(dose = unique(dattab_new$dose)) %>% mutate(logdose = log(dose))
(phat <- predict(object = m, newdat, type="p"))

phat %>% knitr::kable(.,digits = 3)

library(GLMcat)

dattab_new <- dattab_new %>% mutate(logdose = log(dose))
mod_ref_log_c <- glmcat(formula = y0 ~ logdose, ratio = "reference", cdf = "logistic", data = as.data.frame(dattab_new),ref="0",parallel = F)
summary(mod_ref_log_c)
(phat <- predict(object = mod_ref_log_c, newdat, type="prob"))
phat %>% knitr::kable(.,digits = 3)

## (phat <- predict(object = mod_ref_log_c, newdat, type="linear.predictor"))

mod_cum_logis <- glmcat(formula = y0 ~ logdose, ratio = "cumulative", cdf = "logistic", data = as.data.frame(dattab_new),parallel = TRUE)
summary(mod_cum_logis)

(phat <- predict(object = mod_cum_logis, newdat, type="prob"))
phat %>% knitr::kable(.,digits = 3)
```

### Other understanding of the dataset

```{r}
fit0 <- MASS::polr(y0 ~ 1,
                      data = dattab_new,
                      Hess= T)
fit0
#source("https://github.com/rwnahhas/RMPH_Resources/raw/main/Functions_rmph.R")
ilogit <- function(x) exp(x)/(1+exp(x))
ilogit(fit0$zeta)


sf <- function(y) {
  c('Y>=1' = qlogis(mean(y >= 1)),
    'Y>=2' = qlogis(mean(y >= 2)),
    'Y>=3' = qlogis(mean(y >= 3)),
    'Y>=4' = qlogis(mean(y >= 4)),
    'Y>=5' = qlogis(mean(y >= 5)))
}

(s <- with(dattab_new, summary(as.numeric(y0) ~ factor(dose), fun=sf)))


glm(I(as.numeric(y0) >= 2) ~ log(dose), family="binomial", data = dattab_new)
glm(I(as.numeric(y0) >= 3) ~ log(dose), family="binomial", data = dattab_new)
glm(I(as.numeric(y0) >= 4) ~ log(dose), family="binomial", data = dattab_new)

```


### log-logistic with quasi-binomial

- TRUE ER50=12.18.

```{r}
getEC50 <- function(mod){
  if("glm" %in% class(mod)){
    ec50 <- MASS::dose.p(mod,p=0.5)
    se <- attr(ec50,"SE")
    xmin <- as.numeric(ec50 - 1.96*se)
    xmax <- as.numeric(ec50 + 1.96*se) 
    res <- data.frame(EC50=exp(as.numeric(ec50)),
    lower=exp(xmin),upper=exp(xmax),se=se)
  }else{
    if("glmmPQL" %in% class(mod)){
      res1 <- dose.p.glmmPQL(mod,cf=1:2,p=0.5)
      se <- attr(res1,"SE")
      res <- data.frame(EC50=exp(res1[1]),lower=exp(res1[1]-1.96*se),upper=exp(res1[1]+1.96*se),se=se)
    }else{
      res1 <- ED(mod,50,interval="delta")
      res <- data.frame(EC50=res1[1,"Estimate"],lower=res1[1,"Lower"],upper=res1[1,"Upper"],se=res1[1,"Std. Error"])

    }
  }
  return(res)
}

dose.p.glmmPQL <- function(obj,cf=1:2,p=0.5){
  eta <- obj$family$linkfun(p)
  b <- obj$coefficients$fixed[cf]
  x.p <- (eta - b[1L])/b[2L]
  names(x.p) <- paste("p = ", format(p), ":", sep = "")
  pd <- -cbind(1, x.p)/b[2L]
  SE <- sqrt(((pd %*% vcov(obj)[cf, cf]) * pd) %*% c(1, 1))
  res <- structure(x.p, SE = SE, p = p)
  class(res) <- "glm.dose"
  res
}
```


```{r warning=FALSE,message=FALSE}
mod <- glm(yy2~log(dose),data = dattab_new,family = quasibinomial)
summary(mod)
ER50 <- exp(-coef(mod)[1]/coef(mod)[2])
ER50
getEC50(mod)
pred <- predict(mod,newdata = dattab_new,type = "response")
dattab_new$pred <- pred
dattab_new %>% group_by(y0) %>% summarise(n=n(),meany=mean(yt),meanyy2=mean(yy2),meanEst=mean(pred)) %>% knitr::kable(.,digits = 3)

## Consider Replicate effect
modr <- MASS::glmmPQL(yy2~ log(dose),random=~1|rep,family="quasibinomial",data=dattab_new)
ER50 <- exp(-coef(modr)[1]/coef(modr)[2])
exp(-modr$coefficients$fixed[1]/ modr$coefficients$fixed[2])
summary(modr)
getEC50(modr)
pred <- predict(modr,newdata = dattab_new,type = "response")
dattab_new$predr <- pred
dattab_new %>% group_by(y0) %>% summarise(n=n(),meany=mean(yt),meanyy2=mean(yy2),meanEst=mean(pred),meanEstR=mean(predr)) %>% knitr::kable(.,digits = 3)
```


Note that glmer and glmmPQL (based on lme from the nlme pacakge) differs in terms of parameter estimation algorithm and nlme is not optimized for dealing with crossed random effects, which are associated with a sparse design matrix. See more in the book from Pinheiro & Bates.[^1] 
```{r}
uniroot(function(x) {
  y <- as.numeric(predict(modr, newdata = data.frame(dose=x,rep=1),level=0) - 0.5)
  },lower =10, upper = 25)$root
```

### Dose Response for continuous data

```{r warning=FALSE,message=FALSE}
ggplot(dattab_new,aes(x = dose, y=yy2))+geom_point() +  geom_smooth(method="glm", method.args=list(family="quasibinomial"), formula="y ~ log(x)",
                       se =TRUE, size=1.5)+scale_x_log10() +ggtitle(paste("log-logistic + quasibinomial, ER50=",round(ER50,2)))

library(drc)
mod_drc  <- drm(yy2 ~ dose, data = dattab_new, fct = LL.2())
## mod_drc  <- drm(yy2 ~ dose, data = dattab_new, fct = LL.2(),type = "binomial")

summary(mod_drc)
ED(mod_drc, c(50), interval = "delta")

plot(mod_drc, type = "all",main="log-logistic + normal, ER50= 12.74", broken = TRUE)
plot(mod_drc, broken = TRUE, type="confidence", add=TRUE)

```

### DRM Ordinal

```{r}
library(drc)
library(bmd)
dat_ord <- dattab_new %>% group_by(y0,dose) %>% summarise(n=n()) %>% ungroup() %>% pivot_wider(names_from = y0,values_from = n)
dat_ord <- dat_ord %>% mutate(across(c(`0`,`A`,`B`,`C`,`D`,`E`,`F`),~ifelse(is.na(.),0,.))) %>% mutate(total=rowSums(across(c(`0`,`A`,`B`,`C`,`D`,`E`,`F`))))
mod_ord <- drmOrdinal(levels = unique(dattab_new$y0),  weights="total",dose = "dose", data = dat_ord, fct = LL.2())
plot(mod_ord) # uses ggplot

bmdOrdinal(mod_ord, bmr = 0.5, backgType = "modelBased", def = "excess")

```



## GLMcat Basics


```{r}
library(GLMcat)
data("DisturbedDreams")
summary(DisturbedDreams)
head(DisturbedDreams)
```

```{r}
DisturbedDreams$Level <- as.factor(as.character(DisturbedDreams$Level))
mod_ref_log_c <- glmcat(
formula = Level ~ Age, ratio = "reference",
cdf = "logistic", ref_category = "Very.severe",
data = DisturbedDreams
)
```

```{r}
summary(mod_ref_log_c)
```


```{r}
# Random observations
set.seed(13)
ind <- sample(x = 1:nrow(DisturbedDreams), size = 3)
# Probabilities
predict(mod_ref_log_c, newdata = DisturbedDreams[ind, ], type = "prob")
DisturbedDreams[ind, ]
```



```{r}
library(bmd)
## help("bmdOrdinal")
library(tidyverse)
```


```{r}
library(drcData)
data(guthion)
library(drc)

guthionS <- subset(guthion, trt == "S")
guthionS.LL <- drmOrdinal(levels = c("alive", "moribund", "dead"), weights = "total", dose = "dose", data = guthionS, fct = LL.2())
plot(guthionS.LL, xlim = c(15,55)) # uses ggplot

bmdOrdinal(guthionS.LL, bmr = 0.1, backgType = "modelBased", def = "excess")


```

## Ordinal Regression in General

- limited information and pertaintially unequal intervals, which requires a clear understanding of its subtleties and complexities, and careful attentions when implementing regression modelling. 
- Proportional Odds Model and Continuation Ratio Model, are standard regression models looking at ordered categories and transitions between them

- Proportional Odds Model: Coefficient (slopes) remains constant across all categories. 
- Continuation Ratio Model: Model cumulative odds ratios.
- Adjacent Categories Model: useful when neighboring categories influence each other. 

[^1] (Pinheiro & Bates, "Mixed-Effects Models in S and S-PLUS", Springer, 2000, pp. 163. )


