---
title: "Dose Finding Studies and Multiple Contrast Test"
description: "Understanding MCPMod"
author: "Zhenglei Gao"
date: "1/06/2025"
categories:
  - code
  - analysis
---

## Materials and Resouces

- https://github.com/openpharma/DoseFinding/
- [MCP-Mod FAQ](https://openpharma.r-universe.dev/articles/DoseFinding/faq.html)


## Performing MCT

```{r include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(DoseFinding)
  library(drc)
  library(ggplot2)
})

data(IBScovars)
```

```{r}
ggplot(IBScovars, aes(x=dose, y=resp, color=gender)) +
  geom_point(position = position_dodge(width = 0.5)) 
## set random seed to ensure reproducible adj. p-values for multiple contrast test
set.seed(12)

## perform (model based) multiple contrast test
## define candidate dose-response shapes
models <- Mods(linear = NULL, emax = 0.2, quadratic = -0.17,
               doses = c(0, 1, 2, 3, 4))
## plot models
plot(models)
```


```{r}
## fit non-linear emax dose-response model
fitemax <- fitMod(dose, resp, data=IBScovars, model="emax",
                  bnds = c(0.01,5))
## display fitted dose-effect curve
plot(fitemax, CI=TRUE, plotData="meansCI")


## fit non-linear emax dose-response model
fitlogistic <- fitMod(dose, resp, data=IBScovars, model="exponential")
## display fitted dose-effect curve
plot(fitlogistic, CI=TRUE, plotData="meansCI")
```



```{r}
## perform multiple contrast test
MCTtest(dose, resp, IBScovars, models=models,
                addCovars = ~ gender)
```


## Targe Dose / MED



### Differences between TD and ED



```{r}
library(DoseFinding)
library(ggplot2)
## example for creating a "full-model" candidate set placebo response
## and maxEff already fixed in Mods call
doses <- c(0, 10, 25, 50, 100, 150)
fmodels <- Mods(linear = NULL, emax = 25,
logistic = c(50, 10.88111), exponential = 85,
betaMod = rbind(c(0.33, 2.31), c(1.39, 1.39)),
linInt = rbind(c(0, 1, 1, 1, 1),
c(0, 0, 1, 1, 0.8)),
doses=doses, placEff = 0, maxEff = 0.4,
addArgs=list(scal=200))
## calculate doses giving an improvement of 0.3 over placebo
## Target doses 63
TD(fmodels, Delta=0.3)
## discrete version
TD(fmodels, Delta=0.3, TDtype = "discrete", doses=doses)
## doses giving 50% of the maximum effect
DoseFinding::ED(fmodels, p=0.5)
DoseFinding::ED(fmodels, p=0.5, EDtype = "discrete", doses=doses)
plot(fmodels, plotTD = TRUE, Delta = 0.3)
```
## References

- Hothorn, T., Bretz, F., and Westfall, P. (2008). Simultaneous Inference in General Parametric Models, Biometrical Journal, 50, 346–363

- Pinheiro, J. C., Bornkamp, B., Glimm, E. and Bretz, F. (2014) Model-based dose finding under model uncertainty using general parametric models, Statistics in Medicine, 33, 1646–1661
