---
title: "Binomial, Quasibinomial, DRC for proportions"
description: "If all 3 types can model percentage data"
author: "Zhenglei Gao"
date: "12/17/2024"
categories:
  - code
  - stats
  - R
editor_options: 
  chunk_output_type: console
---

```{r include=FALSE}
knitr::opts_chunk$set(warning=FALSE,message=FALSE)
```

In this article, I will try to analyze percentage data as proportions, using quasibinomial and nonlinear fit to the log-logistic cumulative distribution function (CDF) curves. 

Proportional data where the denominator (e.g. maximum possible number of presence for a given response type) is not known can be modeled using a Beta distribution. Binomial-logit-Normal model with observation-level random effects can also be applied to proportions to handle over-dispersion, similarly, quasibinomial models.

For quasibinomial, an additional method is to add subject or individual or treatment group level random effect, that way additional noise is added to the model and can count for the spread of the data in addition to the quasi-binomial variance. However, there is no direct R functions available to calculate all the associated uncertainties, post-processing must be implemented. Therefore, the chance of that becoming a standard approach is very low.

The data is generate in two ways, one is based on the logistic CDF with added noise, the other is just using binomial sampling. For the latter, the sample size information is excluded from the final dataset to mimic the percentage data characteristics. We can also generate over-dispersed binomial data following the steps described [here](https://rdrr.io/cran/fitODBOD/man/GenerateBOD.html) or [here](https://www.r-bloggers.com/2019/05/generating-and-modeling-over-dispersed-binomial-data/)

We are comparing the 3 methods with different data generation scheme.

- Method A: simple quasi binomial
- Method B: drc with normal spread around logistic CDF
- Method C: quasi-binomial with random effect  (`MASS::glmmPQL(percentages/100~ conc,random=~1|conc,family="quasibinomial",data=data)`)

The conclusions are: 

1. For the data generated based on logistic CDF, the standard DRC approach fit the data much better. In terms of EC50 estimation, ...
2. For the data generated using binomial sampling, the 

The data is generated using the cumulative function for logistic distribution. Note that the logistic function is just an inverse logit ($\log(p/(1-p))$) function in R. The code is modified based on [this blog post](https://www.r-bloggers.com/2016/01/s-shaped-data-smoothing-with-quasibinomial-distribution-2/). 
$$F(x;\mu,s) = \frac{1}{2} + \frac{1}{2} \tanh((x-\mu)/2s) = \frac{1}{1+e^{-(x-\mu)/s}}$$

where,  $\tanh(x) = \frac{\sinh(x)}{\cosh(x)} = \frac{e^x - e^{-x}}{e^x + e^{-x}}$, is the [hyperbolic tangent function](https://en.wikipedia.org/wiki/Hyperbolic_functions) that maps real numbers to the range between -1 and 1. 

Quantile of the CDF is then $\mu+s\log(\frac{p}{1-p})$, therefore, the EC50 should be $\mu$ or $\exp(\mu)$.

Random noises are added afterwards to the logistic distribution CDF.

We simulate $n$ studies over concentration $x$, denoted, $X_{1}, X_{2}, …, X_{n}$ for $k$ study $X=(x_{1}, x_{2}, …, x_{k})$, where $k$ is the number of different studies with different $\mu$ and $\sigma$. 

Let’s say there are $k=6$ study groups with the following parameter sets, $\mu = \{9,2,3,5,7,5\}$  and $s=\{2,2,4,3,4,2\}$

```{r}
generate_logit_cdf <- function(mu, s, 
                               sigma_y=0.1, 
                               x=seq(-5,20,0.1)) {
  x_ms <- (x-mu)/s
  y    <- 0.5 + 0.5 * tanh(x_ms)  
  y    <- abs(y + rnorm(length(x), 0, sigma_y))
  ix   <- which(y>=1.0)
  if(length(ix)>=1) { 
    y[ix] <- 1.0
  }
  return(y)
}
tanh(0)
set.seed(424242)
x      <- seq(-5,15,0.025) 
mu_vec <- c(1,2,3,5,7,8)   # 6 Studies
idx <- sapply(mu_vec,function(mu) which(x==mu))
s_vec  <- c(2,2,4,3,4,2)
# Synthetic Study ID
studies_df<- mapply(generate_logit_cdf, 
                              mu_vec, 
                              s_vec, 
                              MoreArgs = list(x=x))
# Give them names
colnames(studies_df) <- c("Study1", "Study2", "Study3", "Study4", "Study5", "Study6")
  
dim(studies_df)
```

```{r fig.width=8,fig.height=5}
library(ggplot2)
theme_set(theme_bw())
library(tidyverse)
df_all <- tidyr::pivot_longer(data.frame(x=(1:length(x))/10,studies_df),cols=2:7)
true_ec50 <- ((1:length(x))/10)[idx]
colnames(df_all) <- c("x", "study", "y")
df_all$study <- as.factor(df_all$study)

p_quasibinomial<- ggplot(df_all, aes(x=x, y=y, color=study)) +
        geom_point(alpha=0.2)  +
        scale_color_viridis_d()+
        # scale_color_brewer(palette = "Reds") + 
        # #geom_smooth(method="loess", se = FALSE, size=1.5) +
        geom_smooth(aes(group=study),method="glm", method.args=list(family="quasibinomial"), formula="y ~ log(x)",
                       se =TRUE, size=1.5) +
        xlab("Concentration") + 
        ylab("% Injury / 100") + ggtitle("Quasibinomial Fit")

p_quasibinomial#+scale_x_log10()

p_quasibinomial +facet_wrap(~study)
```
```{r}
library(drc)
p_drc<- ggplot(df_all, aes(x=x, y=y, color=study)) +
        geom_point(alpha=0.2)  +
        scale_color_viridis_d()+
        # scale_color_brewer(palette = "Reds") + 
        # #geom_smooth(method="loess", se = FALSE, size=1.5) +
        geom_smooth(aes(group=study),method="drm", method.args=list(fct=LL.2()), formula="y ~ x",
                       se = FALSE, size=1.5) +
        xlab("Concentration") + 
        ylab("% Injury / 100") + ggtitle("DRC nonlinear normal fit")
p_drc#+scale_x_log10()

p_drc + facet_wrap(~study)

```

For each study, we can calculate EC50 and confidence intervals.

```{r}
getEC50 <- function(mod){
  if(class(mod)[1]=="glm"){
    ec50 <- MASS::dose.p(mod,p=0.5)
    se <- attr(ec50,"SE")
    xmin <- as.numeric(ec50 - 1.96*se)
    xmax <- as.numeric(ec50 + 1.96*se) 
    res <- data.frame(EC50=exp(as.numeric(ec50)),
    lower=exp(xmin),upper=exp(xmax))
  }else{
    res1 <- ED(mod,50,interval="delta")
    res <- data.frame(EC50=res1[1,"Estimate"],lower=res1[1,"Lower"],upper=res1[1,"Upper"])
  }
  return(res)
}
library(drc)
predictCIglm <- function(mod,dat,newdata=NULL){
 if(is.null(newdata)){
    preddata <- with(dat, data.frame(x = seq(min(x), max(x), length = 100)))
 }else predata <- newdata
  preds <- predict(mod, newdata = preddata, type = "link", se.fit = TRUE)
  critval <- 1.96 ## approx 95% CI qnorm(0.975)
  upr <- preds$fit + (critval * preds$se.fit)
  lwr <- preds$fit - (critval * preds$se.fit)
  fit <- preds$fit
  fit2 <- mod$family$linkinv(fit)
  upr2 <- mod$family$linkinv(upr)
  lwr2 <- mod$family$linkinv(lwr)
  preddata$Prediction <- fit2 
  preddata$Lower <- lwr2 
  preddata$Upper <- upr2 
  return(preddata)
}
preddrc <- function(mod,dat,newdata=NULL){
 if(is.null(newdata)){
    preddata <- with(dat, data.frame(x = seq(min(x), max(x), length = 100)))
 }else predata <- newdata
 preds <- predict(mod,newdata=preddata,interval="confidence")
 preddata <- cbind(preddata,preds)
 return(preddata)
}
modelall <- function(dat){
  mod1 <- glm(y~log(x),data=dat,family = "quasibinomial")
  mod2 <- drm(y~x,fct=LL.2(),data=dat)
  ## mod3 <- MASS::glmmPQL(y ~ log(x),random=~1|x,family="quasibinomial",data=dat) 
  ## Estimation the same as quasibinomial without random effects, uncertainty estimation wider. 
  ## mod3 <- lme4::glmer(y ~ log(x)+ (1|x),family="binomial",data=dat) ## dose not work without sample size.
  # mod4 <- betareg::betareg(y ~ log(x),link="logit",data=dat)
  # y3<- predict(mod3,newdata=data.frame(x=dat$x),type="response")
  y1 <- predictCIglm(mod1,dat=dat)
  y1$model <- "quasibinomial"

  y2 <- preddrc(mod2,dat = dat)
  y2$model <- "drc LL.2"
  preddata <- rbind(y1,y2)
  #ec4 <- uniroot(function(x) predict(mod4, newdata = data.frame(x=x)) - 0.5,lower = 1, upper = 80)$root
  ## 19.55968 beta regression does not perform better than binomial or quasi-binomial regression!
  ec1 <- getEC50(mod1)
  ec2 <- getEC50(mod2)
  ec50 <- data.frame(rbind(ec1,ec2),model=c("quasibinomial","drc LL.2"))
  return(list(ec50=ec50,predata=preddata))
}
```


```{r}
df_nested <- df_all %>% group_by(study) %>% nest()
dfres <- df_nested %>% mutate(modres=map(data,modelall))
ec50 <- dfres %>% mutate(ec50=modres[[1]][1]) %>% dplyr::select(-c(data,modres)) %>% unnest(cols=c(ec50)) %>% ungroup() %>% mutate(TrueEC50 = rep(true_ec50,each=2))
knitr::kable(ec50)
```

```{r}
predres <- dfres %>% mutate(preds=modres[[1]][2]) %>% dplyr::select(-c(data,modres)) %>% unnest(cols=c(preds)) %>% ungroup() 
```

```{r}
p_drc_qb<- ggplot(df_all, aes(x=x, y=y, color=study)) +
        geom_point(alpha=0.1)  +
        scale_color_viridis_d()+
        # scale_color_brewer(palette = "Reds") + 
        # #geom_smooth(method="loess", se = FALSE, size=1.5) +
        geom_smooth(aes(group=study),method="drm", method.args=list(fct=LL.2()), formula="y~x",
                       se = FALSE, size=1.5) +
                       geom_smooth(aes(group=study),method="glm", method.args=list(family="quasibinomial"), formula="y~log(x)",
                       se = FALSE, size=1.5,lty=3)+
        xlab("Concentration") + 
        ylab("% Injury / 100") 
p_drc_qb + facet_wrap(~study)
p_drc_qb <- ggplot(df_all, aes(x=x, y=y)) +
        geom_point(alpha=0.1)  +
        scale_color_viridis_d()+
        # scale_color_brewer(palette = "Reds") + 
        # #geom_smooth(method="loess", se = FALSE, size=1.5) +
        geom_smooth(aes(group=study),method="drm", method.args=list(fct=LL.2()), formula="y~x",
                       se = FALSE, size=1.5,col=scales::hue_pal()(2)[1]) +
                       geom_smooth(aes(group=study),method="glm", method.args=list(family="quasibinomial"), formula="y~log(x)",
                       se = FALSE, size=1.5,lty=3,col=scales::hue_pal()(2)[2])+
        xlab("Concentration") + 
        ylab("% Injury / 100")  + facet_wrap(~study)
 p_drc_qb + geom_hline(yintercept=0.5) + geom_ribbon(data=predres,aes(x=x,ymin=Lower,ymax=Upper,y=Prediction,fill=model),alpha=0.3)



p_drc_qb_beta <- ggplot(df_all, aes(x=x, y=y, color=study)) +
        geom_point(alpha=0.1)  +
        scale_color_viridis_d()+geom_smooth(method="betareg", size=1.5,method.args=list(link="logit"),lty=1,formula="y~log(x)")+ #  geom_smooth(aes(group=study),method="drm", method.args=list(fct=LL.2()), formula="y~x",
 #                      se = FALSE, size=1.5) +
                       geom_smooth(aes(group=study),method="glm", method.args=list(family="quasibinomial"), formula="y~log(x)",
                       se = FALSE, size=1.5,lty=3)+
        xlab("Concentration") + 
        ylab("% Injury / 100") 
p_drc_qb_beta

#  Failed to fit group 5.
#  Caused by error in `x$terms`:
# ! $ operator is invalid for atomic vectors 
# library(MASS)
# ggplot(df_all, aes(x=x, y=y, color=study)) +
#   geom_point(alpha=0.1)  +
#   scale_color_viridis_d() +geom_smooth(aes(group=study),method="glmmPQL", method.args=list(family="quasibinomial",random = ~1|x), formula="y ~ log(x)",
#                                        se = FALSE, size=1.5,lty=1) +
#   geom_smooth(aes(group=study),method="glm", method.args=list(family="quasibinomial"), formula="y~log(x)",
#               se = TRUE, size=1.5,lty=3)+
#   xlab("Concentration") + 
#   ylab("% Injury / 100") 
```

## Data coming from binomial

### Design of data simulation
- 7 treatment groups, including control
- Cover full dose response
- Concentration: `x <- c(0,2^(seq(0.1,4,by=1)))`
- Effect/Response: `p <- plogis(x)`
- 

```{r}
x <- c(2^(seq(-8,7,by=2)))
x
px <- plogis(log(x)) ## inverse logit, I know exactly the true value should be x=1!!!!! (the 5th value)
px
plot(x,px,type="l", log="x")
plogis(0)
```


```{r}
simbinomdata <- purrr::map_dfr(lapply(1:length(x),function(i,n=10,size=100){
  pi <- px[i]
  x1 <- rep(x[i],n)
  yi <- rbinom(n=n,size=size,prob=pi)/size
  return(data.frame(x=x1,y=yi))
}),as.list)

ggplot(simbinomdata,aes(x=x,y=y))+geom_point()+scale_x_log10()+geom_line(data=data.frame(x=x,y=px),aes(x=x,y=px))
```

```{r}
simbinomdata <- simbinomdata |> mutate(size = 100, damagecount= size*y)

mod0 <- glm(cbind(damagecount,size-damagecount)~log(x),data=simbinomdata,family = "binomial")
summary(mod0)
ec50 <- MASS::dose.p(mod0,p=0.5)
## check against MASS does.p function
ecx <- (VGAM::logitlink(0.5)-coef(mod0)[1])/coef(mod0)[2]
pd <- -cbind(1, ecx)/coef(mod0)[2]
ff = as.matrix(vcov(mod0)[1:2,1:2])
se <- sqrt(((pd %*% ff )* pd) %*% c(1, 1))
se
ec50
exp(ec50)
ypred <- predict(mod0,newdata=data.frame(x=x),type="response",se.fit=TRUE)
ggplot(simbinomdata,aes(x=x,y=y))+geom_point()+scale_x_log10()+geom_line(data=data.frame(x=x,y=px),aes(x=x,y=px))+geom_line(data=data.frame(x=x,y=ypred$fit),aes(x=x,y=y),col="red")+geom_ribbon(data=data.frame(x=x,ymin=ypred$fit-1.96*ypred$se.fit,ymax=ypred$fit + 1.96*ypred$se.fit,y=ypred$fit),aes(ymin=ymin,ymax=ymax),fill="red",alpha=0.1)

mod1 <- glm(y~log(x),data=simbinomdata,family = "quasibinomial")
summary(mod1)
ec50_qb <- MASS::dose.p(mod1,p=0.5)
ec50_qb
ec50
ypred1 <- predict(mod1,newdata=data.frame(x=x),type="response",se.fit=TRUE)
p1 <- ggplot(simbinomdata,aes(x=x,y=y))+geom_point()+scale_x_log10()+geom_line(data=data.frame(x=x,y=px),aes(x=x,y=px))+geom_line(data=data.frame(x=x,y=ypred$fit),aes(x=x,y=y),col="red")+ geom_ribbon(data=data.frame(x=x,ymin=ypred1$fit-1.96*ypred1$se.fit,ymax=ypred1$fit + 1.96*ypred1$se.fit,y=ypred$fit),aes(ymin=ymin,ymax=ymax),fill="green",alpha=0.1)+geom_line(data=data.frame(x=x,ymin=ypred1$fit-1.96*ypred1$se.fit,ymax=ypred1$fit + 1.96*ypred1$se.fit,y=ypred$fit),aes(x=x,y=y),col="green")+geom_ribbon(data=data.frame(x=x,ymin=ypred$fit-1.96*ypred$se.fit,ymax=ypred$fit + 1.96*ypred$se.fit,y=ypred$fit),aes(ymin=ymin,ymax=ymax),fill="red",alpha=0.2)
p1
p1+scale_x_log10()+coord_cartesian(xlim=c(0.1,10))

mod3 <-  drm(y ~ x,data=simbinomdata,type="continuous",fct=LL.2())
plot(mod3,type="all",main="DRC fit")
ypred3 <- as.data.frame(predict(mod3,newdata=data.frame(x=x),se.fit=TRUE))


df <- rbind(data.frame(x=x,ymin=ypred$fit-1.96*ypred$se.fit,ymax=ypred$fit + 1.96*ypred$se.fit,y=ypred$fit,model="binomial"),
data.frame(x=x,ymin=ypred$fit-1.96*ypred$se.fit,ymax=ypred$fit + 1.96*ypred$se.fit,y=ypred$fit,model="quasibinomial"),
data.frame(x=x,ymin=ypred3$Prediction-1.96*ypred3$SE,ymax=ypred3$Prediction + 1.96*ypred3$SE,y=ypred3$Prediction,model="drc"))
df$trt <- factor(df$x)
p <- ggplot(data=df,aes(x=model,y=y))+geom_errorbar(aes(ymin=ymin,ymax=ymax,col=model))+facet_grid(~trt,scale="free")+geom_point()
p+coord_flip()
p

getEC50(mod0)
EC50 <- data.frame(upper = (ecx+se*1.96)
,lower =  (ecx-se*1.96))
df1 = data.frame(rbind(getEC50(mod0),
getEC50(mod1),getEC50(mod3)),model=c("binmoial","quasibinomial","drc-LL.2"))

```

## Adding noise to binomial data

- Instead of fixed px, each replicate in every treatment group was generated from px+random noise. 

```{r}
simQuasiBinomdata <- purrr::map_dfr(lapply(1:length(x),function(i,n=30,size=20){
  pi <- px[i]
  pin <- rlnorm(n=n,log(px[i]),sd=0.1)
  pin[pin>1] <- 1
  x1 <- rep(x[i],n)
  yi <- rbinom(n=n,size=size,prob=pin)/size
  return(data.frame(x=x1,y=yi))
}),as.list)

simQuasiBinomdata$size <- 20
ggplot(simQuasiBinomdata,aes(x=x,y=y))+geom_point()+scale_x_log10()+geom_line(data=data.frame(x=x,y=px),aes(x=x,y=px))
comp3models <- function(simdata,size=10,...){
  simdata <- simdata |> mutate(damagecount=size*y)
  mod0 <- glm(cbind(damagecount,size-damagecount)~log(x),data=simdata,family = "binomial")
  x <- c(2^(seq(-8,7,by=0.5)))
  ypred <- predict(mod0,newdata=data.frame(x=x),  type="response",se.fit=TRUE)
  
  mod1 <- glm(y~log(x),data=simdata,family = "quasibinomial")
  ypred1 <- predict(mod1,newdata=data.frame(x=x),type="response",se.fit=TRUE)
  
  mod3 <-  drm(y ~ x,data=simdata,type="continuous",fct=LL.2())
  ypred3 <- as.data.frame(predict(mod3,newdata=data.frame(x=x),se.fit=TRUE)) 
  df <- rbind(data.frame(x=x,ymin=ypred$fit-1.96*ypred$se.fit,ymax=ypred$fit + 1.96*ypred$se.fit,y=ypred$fit,model="binomial"),
  data.frame(x=x,ymin=ypred$fit-1.96*ypred$se.fit,ymax=ypred$fit + 1.96*ypred$se.fit,y=ypred$fit,model="quasibinomial"),
  data.frame(x=x,ymin=ypred3$Prediction-1.96*ypred3$SE,ymax=ypred3$Prediction + 1.96*ypred3$SE,y=ypred3$Prediction,model="drc"))
  df$trt <- factor(df$x)
  df1 = data.frame(rbind(getEC50(mod0),
  getEC50(mod1),getEC50(mod3)),model=c("binmoial","quasibinomial","drc-LL.2"))
  df1$Prediction <- c(predict(mod0,data.frame(x=df1$EC50[1]),type="response"),predict(mod1,data.frame(x=df1$EC50[2]),type="response"),predict(mod3,data.frame(x=df1$EC50[3])))
  return(list(pred=df,ec50=df1))
}

res <- comp3models(simQuasiBinomdata,size=20)

p1 <- ggplot(simQuasiBinomdata,aes(x=x,y=y))+geom_point()+scale_x_log10()+geom_line(data=data.frame(x=x,y=px),aes(x=x,y=px))
p2 <- p1 + geom_ribbon(data=res$pred,aes(ymin=ymin,ymax=ymax,col=model,fill=model),alpha=0.1) + geom_line(data=res$pred,aes(col=model))
p2
p2+scale_x_log10()+coord_cartesian(xlim=c(0.1,10))
res$ec50

```


! need to be checked!
```{r}
simfun <- function(x=2^(seq(-8,7,by=2)) ,n=10,size=10,sd=0.1){
  px <- plogis(x)
  simdata <- purrr::map_dfr(lapply(1:length(x),function(i,nn,size,sd){
  pxi <- px[i]
  pin <- rlnorm(n=nn,log(pxi),sd=sd)
  pin[pin>1] <- 1
  x1 <- rep(x[i],n)
  yi <- rbinom(n=nn,size=size,prob=pin)/size
  return(data.frame(x=x1,y=yi))
  },nn=n,size=size,sd=sd),as.list)
  return(simdata)
}

simdata <- simfun(x=2^(seq(-8,7,by=2)),10,10,0.1)
res <- comp3models(simQuasiBinomdata,size=10)
```

```{r}
library(SimEngine)

```



## Numerical Considerations

The quasi-binomial isn't necessarily a particular distribution. The only assumption is that for a generalized linear model with binomial assumption of "p" (mean of the binomial should be $np$), the variance is $\phi$ times the variance for that binomial = np(1-p). In case of missing sample size, then it is just  $\phi p(1-p)$ 

Quasi-binomial is exactly the same as binomial for the coefficient / parameter estimates but not for the standard errors. In fact, quasi-likelihood models do a post-fitting adjustment to the standard errors of the parameters and the associated statistics; they don't change anything about the way the model is fitted. Some analysis uses observation-level random effects (OLRE) rather than quasi-likelihood to handle overdispersion at the level of individual observations. The way to adjust the coefficient standard error and to model OLRE is in [this link](https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#fitting-models-with-overdispersion)

In terms of GLM, the dispersion for non-over-dispersed model like binomial and Poisson is 1. For quasi-families, it is estimated by the residual Chi-squared statistic divided by the residual degrees of freedom. 

For estimation of EC50 or other ECx, other than inverse regression methods such as [here](https://cran.r-project.org/web/packages/investr/) and [here]()

$$\log \Big( \frac{p}{1-p} \Big) = \beta_0 + \beta_1 x.$$

Setting $p = \tfrac{1}{2} = 0.5 $ gives the corresponding explanatory value (here the LC50 concentration):

$$ \mbox{LC}_{50} \equiv \frac{\log(\tfrac{1}{2}/\tfrac{1}{2}) - \beta_0}{\beta_1}
= \frac{\log(1) - \beta_0}{\beta_1}
= - \frac{\beta_0}{\beta_1}.$$

There is a **{dose.p}** function in the MASS package for calculating LCx or ECx together with the standard errors which can be used for the response confidence intervals. The **vcov** function can be used to extract the covariance function and used for estimation the CI of LCx or ECx. An alternative approach is to use bootstrap confidence intervals. 


Since the goal of the data analysis is to measure the association between a set of regression parameters and the outcome, quasibinomial models can be useful with the variance assumption being relaxed. However, it is required to estimate the uncertainty associated with that relationship. 

```{r}
## install.packages("VGAM")
## data:
Concentration <- as.numeric(c(0, 100, 200, 300, 400, 500, 600, 700, 800))
Survival <- as.integer(c(20, 20, 20, 18, 13, 8, 3, 0, 0))
Total <- as.integer(c(20, 20, 20, 20, 20, 20, 20, 20, 20))
Data <- data.frame(Concentration, Survival, Total)
Data$Dead <- Data$Total-Data$Survival
Data$Proport_Surv <- (Data$Total - Data$Dead) / Data$Total

mod0 <- glm(cbind(Survival, Dead) ~ Concentration, Data, family=binomial(link = "logit"))
mod <- glm(cbind(Survival, Dead) ~ Concentration, Data, family=quasibinomial(link = "logit"))

NewData1 <- expand.grid(Conc  = seq(0, 800, length = 9))

P_logit <- predict(mod, newdata = NewData1, se = TRUE, type = "link")
NewData1$P_logit <- exp(P_logit$fit) / (1 + exp(P_logit$fit))
NewData1$SeUp <- exp(P_logit$fit + 1.96*P_logit$se.fit) / (1 + exp(P_logit$fit + 1.96*P_logit$se.fit))
NewData1$SeLo <- exp(P_logit$fit - 1.96*P_logit$se.fit) / (1 + exp(P_logit$fit - 1.96*P_logit$se.fit))
plot <- plot + geom_point(data = Data, aes(y = Proport_Surv, x = Concentration), shape = 1, size = 2.5)
plot <- ggplot() + xlab("Concentration") + ylab("Proportion of surviving organisms") + theme(text = element_text(size=15)) + theme_bw()+ geom_line(data = NewData1, aes(x = Concentration, y = P_logit), colour = "black")+ geom_ribbon(data = NewData1, aes(x = Concentration, ymax = SeUp, ymin = SeLo), alpha = 0.2)+ geom_hline(yintercept = 0.5, linetype = "dashed") + annotate("point", x = 460.1353, y = 0.5, size = 3.25, colour = "blue")

ec = 0.5
library(VGAM)
eta <- VGAM::logitlink(ec) 
beta <- coef(mod)[1:2] 
ecx <- (eta - beta[1])/beta[2] 
pd <- -cbind(1, ecx)/beta[2] 
ff = as.matrix(vcov(mod)[1:2,1:2])
se <- sqrt(((pd %*% ff )* pd) %*% c(1, 1))
upper = (ecx+se*1.96)
lower = (ecx-se*1.96)
df1 = data.frame(ecx, lower, upper)

ecx0 <- (eta - coef(mod0)[1])/coef(mod0)[2] 
ff0 = as.matrix(vcov(mod0)[1:2,1:2])
se0 <- sqrt(((pd %*% ff0 )* pd) %*% c(1, 1))
upper0 = (ecx+se0*1.96)
lower0 = (ecx-se0*1.96)
df0 = data.frame(ecx, lower0, upper0)

plot <- plot + geom_vline(xintercept = df1$ecx, linetype = "dashed")
plot <- plot + geom_vline(xintercept = df1$lower, linetype = "dashed")
plot <- plot + geom_vline(xintercept = df1$upper, linetype = "dashed")
plot
```

For binomial data, the quasibinomial fit generated a narrower CI compared to binomial fit, unexpected. 

```{r}
df0
df1
```

In fact, a limitation of these models is that they cannot yield accurate prediction intervals, the Pearson residuals cannot tell you much about how accurate the mean model is, and information criteria like the AIC or BIC cannot effectively compare these models to other types of models.

```{r}
AIC(mod0)
# [1] 19.40066
try(AIC(mod1))
# > AIC(mod1)
# Error in h(simpleError(msg, call)) :
#   error in evaluating the argument 'object' in selecting a method for function 'AIC': object 'mod1' not found
```




## Generate data 

```{r}
logit_scale = seq(-6, 6, by = 0.1)

plot( logit_scale,plogis(logit_scale), ylab = 'probabilities', type = 'n',main="Logistic Function")
lines( logit_scale,plogis(logit_scale))

plot( 1/(1+exp(-logit_scale)),plogis(logit_scale), ylab = 'probabilities', type = 'l',main="Linear Transformation of above")
```

```{r}
library('MASS')
library('stats')

gen_data <- function(n=100, p=3) {

  set.seed(1)  
  weights <- stats::rgamma(n=n, shape=rep(1, n), rate=rep(1, n))
  y <- stats::rbinom(n=n, size=1, prob=0.5)
  theta <- stats::rnorm(n=p, mean=0, sd=1)
  means <- colMeans(as.matrix(y) %*% theta)
  x <- MASS::mvrnorm(n=n, means, diag(1, p, p))

  return(list(x=x, y=y, weights=weights, theta=theta))  
}

fit_glm <- function(family) {
  data <- gen_data()
  fit <- stats::glm.fit(x = data$x,
                        y = data$y,
                        weights = data$weights,
                        family = family)
  return(fit)
}

fit1 <- fit_glm(family=stats::binomial(link = "logit"))
fit2 <- fit_glm(family=stats::quasibinomial(link = "logit"))

all(fit1$coefficients == fit2$coefficients)



```


```{r}
data1 <- gen_data()
```
```{r}
summary1 <- stats::summary.glm(fit1)
summary2 <- stats::summary.glm(fit2)

print("Equality of unscaled variance-covariance-matrix:")
all(summary1$cov.unscaled == summary2$cov.unscaled)

print("Equality of variance-covariance matrix scaled by `dispersion`:")
all(summary1$cov.scaled == summary2$cov.scaled)

print(summary1$coefficients)
print(summary2$coefficients)
```


## References: Quasi Models
- [Quasipoisson, GLMM, GEE](https://stats.stackexchange.com/questions/201903/how-to-deal-with-overdispersion-in-poisson-regression-quasi-likelihood-negativ/332250#332250)
- [Tutorial for count data](https://online.stat.psu.edu/stat504/lesson/9/9.2-0)
- [Details on prediction CI with bootstrap and inverse link function](https://stackoverflow.com/questions/14423325/confidence-intervals-for-predictions-from-logistic-regression)
