---
title: "GEE and categorical responses"
description: "Try using GEE for categorical responses"
author: "Zhenglei Gao"
date: "01/14/2025"
categories:
  - code
  - analysis
editor_options: 
  chunk_output_type: console
---

## Use the data from previous posts


## Multinomial or ordinal GEE

Data: Rheumatoid self-assessment scores for 302 patients, measured on a five-level ordinal response scale at three follow-up times.

```{r}
library(multgee)
data(arthritis)
intrinsic.pars(y, arthritis, id, time) ## assess the underlying association pattern.
fitmod <- ordLORgee(formula = y ~ factor(time) + factor(trt) + factor(baseline),
  data = arthritis, id = id, repeated = time, LORstr = "uniform")
summary(fitmod)
confint(fitmod)
library(ggstats)
ggcoef_multinom(fitmod)
```

What is the difference between Ordinal GEE and ordinal logistic regression? 

```{r}

```

## GEE Basics

- It is a quasilikelihood approach.
- Could include cluster/group correlation structure(e.g., repeated measurements from one subject)
- No distributional assumption.
- Could handle binary or categorical responses (multinomial).
- Computationally simpler for categorical compared to maximum likelihood approach.
- No likelihood based post hoc testing, comparison or inference.

```{r}
URL <- "http://static.lib.virginia.edu/statlab/materials/data/depression.csv"
dat <- read.csv(URL, stringsAsFactors = TRUE)
dat$id <- factor(dat$id)
dat$drug <- relevel(dat$drug, ref = "standard")
head(dat, n = 3)
library(tidyverse)
with(dat, tapply(depression, list(diagnose, drug, time), mean)) %>% 
  ftable() %>% 
  round(2)

```

```{r}
##install.packages("gee")
library(gee) # version 4.13-25
dep_gee <- gee(depression ~ diagnose + drug*time,
               data = dat, 
               id = id, 
               family = binomial,
               corstr = "independence")
summary(dep_gee)

res <- glm(depression ~ diagnose + drug*time,
               data = dat,
               family = quasibinomial)
summary(res)
```

The “Estimated Scale Parameter” is reported as 0.9854113 = the dispersion parameter = similar to what is reported (Dispersion parameter for quasibinomial family taken to be 0.9854187) when running glm() with family = quasibinomial in order to model over-dispersion.


```{r}
dep_gee2 <- gee(depression ~ diagnose + drug*time,
               data = dat, 
               id = id, 
               family = binomial,
               corstr = "exchangeable")
summary(dep_gee2)


dep_gee3 <- gee(depression ~ diagnose + drug*time,
               data = dat, 
               id = id, 
               family = binomial,
               corstr = "AR-M", Mv = 1)
dep_gee3$working.correlation
```


```{r}
library(lme4) # version 1.1-31
dep_glmer <- glmer(depression ~ diagnose + drug*time + (1|id), 
               data = dat, family = binomial)
summary(dep_glmer, corr = FALSE)

library(ggeffects) # version 1.2.0
plot(ggemmeans(dep_glmer, terms = c("time", "drug"), 
               condition = c(diagnose = "severe"))) + 
  ggplot2::ggtitle("GLMER Effect plot")

```


```{r}
library(emmeans) # version 1.8.4-1
emm_out <- emmeans(dep_gee2, specs = c("time", "drug"), 
        at = list(diagnose = "severe"), 
        cov.keep = "time", 
        regrid = "response") %>% 
  as.data.frame()
library(ggplot2) # version 3.4.1
ggplot(emm_out) +
  aes(x = time, y = prob, color = drug, fill = drug) +
  geom_line() +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL, 
                  color = NULL), alpha = 0.15) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1", guide = NULL) +
  scale_y_continuous(labels = scales::percent) +
  theme_ggeffects() +
  labs(title = "GEE Effect plot", y = "depression")
```



```{r}

library(geepack) # version 1.3.9
dep_gee4 <- geeglm(depression ~ diagnose + drug*time,
                   data = dat,
                   id = id,
                   family = binomial,
                   corstr = "exchangeable")
# create effect plot
plot(ggemmeans(dep_gee4, terms = c("time", "drug"),
               condition = c(diagnose = "severe"))) + 
  ggplot2::ggtitle("GEE Effect plot")
```


## References

- [Getting started with GEE](https://library.virginia.edu/data/articles/getting-started-with-generalized-estimating-equations)
- [geepack R package]()
- [A note on Weights in statistics](https://notstatschat.rbind.io/2020/08/04/weights-in-statistics/)

## Additional references on survey data

- [survey analysis](https://ladal.edu.au/surveys.html)

