---
title: "Hessian and Likelihood"
description: "A discussion on confidence intervals"
author: "Zhenglei Gao"
date: "11/20/2024"
categories:
  - code
  - Stats
  - R
---

I was discussing with Johannes on how to calculate confidence intervals for a growth model, or more in general, confidence interval, credibal interval, prediction interval, tolerance interval. I need to write that up too. I have written him a formula for deriving the SEs from the Hessian matrix of the optim output or using the numDeriv::hessian function. However, our beloved myGenAssist does not agree with me. After a clarification with the assistant, we reached an agreement. 


Anyway, I first did a search on Hessian, likelihood, Fisher information matrix and came across [this post] (https://stats.stackexchange.com/questions/381984/standard-error-from-hessian-matrix-when-likelihood-is-used-rather-than-ln-l). Below is to reproduce the results of that post. 

Denote $$l(x) = \ln L(x)$$ and ' for differentiation,

$$l'(x) = \frac{L'(x)}{L(x)}$$

and

$$l''(x) = \frac{L''(x)}{L(x)}  - \left(\frac{L'(x)}{L(x)}\right)^2$$

At a critical point $x_0$, $L'(x_0)$ is 0, so

$$l''(x_0) = \frac{L''(x_0)}{L(x_0)}$$


The problem seems to be round-off error or *objective functions being much, much smaller than the default stopping-rule thresholds* as WarrenWeckesser and JimB pointed out. 

```{r}
  x = c(1, 3, 4, 5, 7.5, 10, 12, 23, 39, 40)
    
    # The *exact* MLE for the normal distribution is
    #    mu = mean(x),  sigma = sqrt(var(x)*(length(x) - 1)/length(x))
    
    p0 = c(mean=mean(x), sd=sd(x))
    
    Llikelihood <- function(par, x) {
       return(sum(dnorm(x, mean=par["mean"], sd=abs(par["sd"]), log = TRUE)))
    }
    o1 <- optim(par=p0, fn=Llikelihood, x=x, hessian=TRUE, method="CG", control=list(fnscale=-1, reltol=1e-12, maxit=10000))
    
    cat("\n*** Using log-likelihood ***\n\n")
    cat("Estimated parameters:\n")
    o1$par
    cat("\nEstimated Hessian:\n")
    o1$hessian
    cat("\nEstimated standard errors:\n")
    sqrt(diag(solve(-o1$hessian)))
    
    cat("\n*** Using (plain) likelihood ***\n\n")
    
    likelihood <- function(par, x) {
     return(prod(dnorm(x, mean=par["mean"], sd=abs(par["sd"]), log = FALSE)))
    }
    o2 <- optim(par=p0, fn=likelihood, x=x, hessian=TRUE, method="CG", control=list(fnscale=-likelihood(p0, x), reltol=1e-12, maxit=10000))
    
    cat("Estimated parameters:\n")
    o2$par
    cat("\nEstimated Hessian of neg. log likelihood:\n")
    o2$hessian/o2$value
    cat("\nEstimated standard errors:\n")
    sqrt(diag(solve(-o2$hessian/o2$value)))
```


  

```{r}
set.seed(12345)
x = rnorm(10,3,4)
x = rnorm(100,3,4)
Llikelihood <- function(par, x) {
    return(sum(dnorm(x, mean=par["mean"], sd=abs(par["sd"]), log = TRUE)))
 }
o <- optim(par=c(mean=3, sd=4), fn=Llikelihood, x=x, hessian = TRUE, control=list(fnscale=-1))
sqrt(diag(solve(-o$hessian)))


likelihood <- function(par, x) {
  return(prod(dnorm(x, mean=par["mean"], sd=abs(par["sd"]), log = FALSE)))
}

likelihood_amplified <- function(par, x) {
  return(10^128*prod(dnorm(x, mean=par["mean"], sd=abs(par["sd"]), log = FALSE)))
 }
o <- optim(par=c(mean=3, sd=4), fn=likelihood, x=x, hessian = TRUE, control=list(fnscale=-1))
sqrt(diag(solve(-o$hessian/o$value)))

o <- optim(par=c(mean=3, sd=4), fn=likelihood_amplified, x=x, hessian = TRUE, control=list(fnscale=-1))
sqrt(diag(solve(-o$hessian/o$value)))

```
### Other resources

- https://biostatsr.blogspot.com/2017/05/standard-error-from-hessian-matrix-when.html
- https://web.stanford.edu/~mrosenfe/soc_meth_proj3/matrix_OLS_NYU_notes.pdf
- https://uu.diva-portal.org/smash/get/diva2:1810474/FULLTEXT01.pdf
- https://stats.stackexchange.com/questions/52829/why-is-a-likelihood-ratio-test-distributed-chi-squared
- https://books.google.fr/books?id=vmoFw0pvS1sC&printsec=frontcover&hl=fr&source=gbs_ge_summary_r&cad=0#v=onepage&q&f=false


