---
title: "Cluster Analysis Refresher"
description: "Simple cluster analysis"
author: "Zhenglei Gao"
date: "02/18/2025"
categories:
  - code
  - stats
  - R
editor_options: 
  chunk_output_type: console
---

::: callout-warning
To be written.
:::


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

I started working on a project clustering a trait database so that I can get up to 15 clusters of species groups. Hence I am refreshing on the potential algorithms I could use. 

## Algorithms

There are different types of clustering alogrithm with different distance measures, including density-based, distribution-based, centroid-based, hierarchical-based, model-based. 

- The common 4 distances measures are Enclidaen, Manhattan, Correlation and Eisen. 

We focus on a few for the moment. 

- Partitioning Clustering 
    - K-means clustering (centroid-based)
    - K-Medoids (PAM)
    - CLARA
- DBSCAN (density-based spatial clustering of applications with noise)
- Hierarchical Clustering
    - - Agglomerative Hierarchy clustering algorithm    
- Spectral Clustering


Cluster GoF (clValid)

- Internal: the connectivity, the silhouette coefficient and the Dunn index 
- Stability: average proportion of non-overlap, average distance, average distance between means, figure of metrit.

## CLARA

CLARA is used in another python ML project. Compared to k-means clustering, CLARA is an extention of k-medoids methods to deal with large data. It is achieved by iterative sampling and then clusterting. 

## Using tidymodels or tidy pipeline


## DBSCAN




## Agglomerative hierarchy clustering

```{r}
library(cluster)
data(plantTraits)

## Calculation of a dissimilarity matrix

dai.b <- daisy(plantTraits,
               type = list(ordratio = 4:11, symm = 12:13, asymm = 14:31))

## Hierarchical classification
agn.trts <- agnes(dai.b, method="ward")
plot(agn.trts, which.plots = 2, cex= 0.6)
plot(agn.trts, which.plots = 1)
cutree6 <- cutree(agn.trts, k=6)
cutree6

## Principal Coordinate Analysis
cmdsdai.b <- cmdscale(dai.b, k=6)
plot(cmdsdai.b[, 1:2], asp = 1, col = cutree6)
```

