---
title: "Preparing A Gental Intro to GLMM"
description: "Audience are ecotoxicologists"
author: "Zhenglei Gao"
date: "09/25/2025"
categories:
  - code
  - R
execute:
  enabled: true
editor_options: 
  chunk_output_type: console
---


During the preparation, I ask Grok and Gemini to help me prepare some examples. Here is what I used as prompts for one of the example. 

The simulation design: 

- ...

In the visualization plot, I want to: 
- Overlay GLM lines (no random effects) versus GLMM field-specific prediction lines (with random intercepts included).
- Test the Control vs Treated effect under GLM and under GLMM via likelihood-ratio tests.
- Estimate statistical power for detecting the treatment effect via simulation for GLM vs GLMM.

I got some very good code snippets but the results are not exactly what I was looking for, I have to ajust here and there as usual. LLM still can improve for these kind of tasks. Especially like envrionment set up, small typos like `loglik.p` instead `loglik`. In these scenarios, I am telling LLM what I want in the end, and LLM should work like an efficient junior colleauge who can understand my requirements and can either code from scratch or find the right function in the right package to shorten the code. My experience is that LLM usually meet me in the middle. They often utilize popular packages, but not always find the most efficient solution. For example, for hierarchical data simulation, there are already some R packages can do this task by a single function. LLM tends to code this part by themselves. It is good because I think for people reading the code, it is actually easy to understand what is intended compared to reading help files of a function written by someone else. 






## Model Diagnostics


- I would use “assumption checks and model diagnostics” or “model adequacy checks” rather than “validation.” The poster labels this step “Validation” and illustrates it with model validation plots, but in practice these are diagnostic checks of assumptions and fit. 

What to check for GLMM (Negative Binomial with log link)

- Residual uniformity and overall fit via simulated residuals.
- Overdispersion.
- Zero inflation.
- Temporal autocorrelation (sessions are ordered).
- Random-effect structure (e.g., normality of field intercepts).
- Observed vs predicted patterns (by session and treatment).


# GAMM and NLMM

Here’s a concise “Nonlinear models” block you can add, with conceptual expressions and ecotoxicology examples. Each formula is wrapped with$.

Concepts
- NLMM (Nonlinear Mixed-Effects Model, Gaussian errors)
  - $y_{ij} = f(x_{ij}; \boldsymbol{\theta}) + u_i + \epsilon_{ij}$
  - $u_i \sim \mathcal{N}(0, \sigma_u^2), \quad \epsilon_{ij} \sim \mathcal{N}(0, \sigma^2)$
  - f(·) is nonlinear in parameters/predictors; random effects enter additively on the observation scale.

- GNLMM (Generalized Nonlinear Mixed-Effects Model, non-Gaussian)
  - $y_{ij} \sim \text{Distribution}(\mu_{ij}, \phi)$
  - $g(\mu_{ij}) = h(x_{ij}; \boldsymbol{\theta}) + u_i$
  - h(·) is a nonlinear function; random effects enter additively on the link scale.

Ecotoxicology examples
- NLMM: Body weight following Gompertz growth, treatment affects growth rate; field-specific asymptote
  - $w_{ij} = A_i \, \exp\{-\exp[-k(\text{Treatment}_i)\, (t_j - t_0)]\} + \epsilon_{ij}$
  - $A_i = A + u_i, \quad u_i \sim \mathcal{N}(0, \sigma_A^2), \quad \epsilon_{ij} \sim \mathcal{N}(0, \sigma^2)$
  - $k(\text{Treatment}_i) = k_C + \beta_T \, \mathbb{I}(\text{Treatment}_i=\text{Treated})$

- GNLMM: Vole abundance (MNA) with mechanistic population dynamics and treatment effect; Negative Binomial likelihood
  - Discrete growth with treatment-induced decay:
    - $\mu_{ij} = N_{0,i} \, \exp\big(r \, t_j\big) \, \exp\big(-\delta \, \mathbb{I}(\text{Treatment}_i=\text{Treated}) \, t_j\big)$
    - $\text{MNA}_{ij} \sim \text{NB}(\mu_{ij}, k)$
    - $\log N_{0,i} = \log N_0 + u_i, \quad u_i \sim \mathcal{N}(0, \sigma_{N_0}^2)$
  - Alternatively, a density-dependent form (Beverton–Holt) with random carrying capacity:
    - $\mu_{i,j} = \frac{K_i \, R \, \mu_{i,j-1}}{1 + \alpha \, \mu_{i,j-1}} \, \exp\big(-\delta \, \mathbb{I}(\text{Treatment}_i=\text{Treated})\big)$
    - $\text{MNA}_{i,j} \sim \text{NB}(\mu_{i,j}, k), \quad \log K_i = \log K + u_i, \; u_i \sim \mathcal{N}(0, \sigma_K^2)$

- GNLMM: Proportion reproductively active with an Emax dose-response; Binomial likelihood
  - $\text{logit}(p_{ij}) = \text{logit}(p_0) + \frac{E_{\max} \, \text{dose}_{ij}}{EC_{50} + \text{dose}_{ij}} + u_i$
  - $y_{ij} \sim \text{Binomial}(n_{ij}, p_{ij}), \quad u_i \sim \mathcal{N}(0, \sigma_u^2)$

When to use

- Choose NLMM/GNLMM when biology/mechanism implies nonlinear relationships (growth curves, dose-response, population dynamics), and you need to:
  - Respect non-Gaussian endpoints (counts, proportions) via appropriate links/distributions (GNLMM).
  - Capture hierarchical/repeated-measures designs via random effects (field/plot/animal).


  ## RM-ANOVA vs. (G)LMM



**When p‑values tend to be the same**

- Balanced design (equal numbers per group/subject/time) with no missing data.
- Time modeled as a categorical factor.
- Random intercept only (correlation structure equivalent to compound symmetry).
- Homoscedastic residuals (equal variances across groups/times).
- Consistent contrasts and the same hypothesis type (e.g., Type III tests in both).
- Gaussian response with identity link (no transformation or non-Gaussian likelihood).

**When they differ**

- Unbalanced or incomplete data (missing sessions/subjects). LMMs handle unbalance; RM‑ANOVA does not gracefully.
- Sphericity violated. RM‑ANOVA needs GG/HF corrections (inflates p); LMM can model correlation directly (random slopes, AR(1), etc.), often yielding different p‑values.
- Heterogeneous variances across groups/times. LMMs can model variance heterogeneity; RM‑ANOVA assumes equality.
- More complex random effects (random slopes, crossed/random effects beyond subject intercepts). RM‑ANOVA can’t represent these; LMM can.
- Non‑Gaussian endpoints (counts, proportions). GLMM vs RM‑ANOVA will diverge by design.
- Different SS types or contrast coding (Type I vs II vs III, treatment vs sum contrasts) lead to different tests.
- Small samples and df approximations (Satterthwaite/Kenward‑Roger in LMM) can shift p‑values relative to classical ANOVA dfs.
- Using ML vs REML for comparing nested mixed models can change inference; RM‑ANOVA doesn’t have this distinction.


### When they are the same
  Below is a self-contained R example that:
- Simulates a simple avian body weight dataset with repeated measures in a hierarchical design (birds nested in treatment, measured over time).
- Compares a repeated-measures ANOVA with a linear mixed model (LMM) for treatment effects (also shows time and interaction).
- Plots the data (spaghetti by bird + group means).
- Produces a compact outcome table comparing p-values from RM-ANOVA vs LMM.


```{r}

# Packages
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(lme4)
  library(lmerTest)  # p-values for lmer
})

set.seed(123)

# 1) Simulate hierarchical repeated-measures data
simulate_body_weight <- function(
  n_groups = 4,                        # Control + 3 treatments
  group_names = c("Control","T1","T2","T3"),
  n_birds_per_group = 12,
  n_time = 6,                          # repeated measures
  beta0 = 20,                          # baseline weight (g)
  treat_effects = c(0, 1.0, -0.5, -1.5), # group effects (Control=0)
  time_slope = 0.8,                    # growth per time (g)
  treat_time_slopes = c(0, 0.2, -0.1, -0.3), # treatment-specific deviations in slope
  sd_id = 1.5,                         # random intercept SD (bird-level)
  sd_resid = 1.0                       # residual SD
) {
  stopifnot(n_groups == length(group_names), length(treat_effects) == n_groups)
  ids <- unlist(lapply(seq_len(n_groups), function(g) {
    paste0(group_names[g], "_", seq_len(n_birds_per_group))
  }))
  df <- expand.grid(
    id = ids,
    time = seq_len(n_time),
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  )
  # Map id -> treatment
  id_map <- data.frame(
    id = ids,
    treatment = rep(group_names, each = n_birds_per_group),
    stringsAsFactors = FALSE
  )
  df <- df %>% left_join(id_map, by = "id")
  df$treatment <- factor(df$treatment, levels = group_names)
  
  # Center time for stability; keep both factor and centered numeric
  df <- df %>%
    mutate(
      time_f = factor(time),
      time_c = time - mean(time)
    )
  
  # Random intercept per bird
  re_id <- rnorm(length(unique(df$id)), 0, sd_id)
  names(re_id) <- unique(df$id)
  
  # Fixed effects by treatment and time (allow treatment-specific slopes)
  treat_idx <- match(df$treatment, group_names)
  tr_eff <- treat_effects[treat_idx]
  tr_slope <- time_slope + treat_time_slopes[treat_idx]
  
  mu <- beta0 + tr_eff + tr_slope * df$time_c + re_id[df$id]
  y <- rnorm(nrow(df), mean = mu, sd = sd_resid)
  
  df$weight <- y
  df
}

dat <- simulate_body_weight()

# 2) Repeated-measures ANOVA (between: treatment; within: time)
#    Classic RM-ANOVA requires within-subject (id) Error term
aov_rm <- aov(weight ~ treatment * time_f + Error(id/time_f), data = dat)
sum_aov <- summary(aov_rm)

# 3) Linear Mixed Model (random intercept for id)
#    Keep time as factor to align with RM-ANOVA fixed-effects structure
options(contrasts = c("contr.sum", "contr.poly"))  # for type III
m_lmm <- lmer(weight ~ treatment * time_f + (1 | id), data = dat)
anova_lmm <- anova(m_lmm, type = 3)  # Satterthwaite df from lmerTest

# 4) Extract comparable results (Treatment, Time, Treatment:Time)
# RM-ANOVA tables are in different Error strata:
# - Treatment (between) in "Error: id"
# - Time and Treatment:Time in "Error: id:time_f"

get_aov_p <- function(sum_list, stratum, term) {
  # Navigate summary(aov) object
  tab <- sum_list[[stratum]][[1]]
  # Handle row names carefully (factor contrasts appear)
  rn <- rownames(tab)
  hit <- which(startsWith(rn, term))
  if (length(hit) == 0) return(NA_real_)
  as.numeric(tab[hit[1], "Pr(>F)"])
}

p_aov_treat <- get_aov_p(sum_aov, "Error: id", "treatment")
p_aov_time  <- get_aov_p(sum_aov, "Error: id:time_f", "time_f")
p_aov_int   <- get_aov_p(sum_aov, "Error: id:time_f", "treatment:time_f")

extract_lmm_row <- function(anova_tab, term_label) {
  # anova_lmm rownames should match "treatment", "time_f", "treatment:time_f"
  row <- anova_tab[term_label, , drop = FALSE]
  list(F = as.numeric(row$`F value`), df = as.numeric(row$DenDF), p = as.numeric(row$`Pr(>F)`))
}

row_treat <- extract_lmm_row(anova_lmm, "treatment")
row_time  <- extract_lmm_row(anova_lmm, "time_f")
row_int   <- extract_lmm_row(anova_lmm, "treatment:time_f")

# 5) Outcome comparison table
outcome <- tibble::tibble(
  Effect = c("Treatment", "Time", "Treatment × Time"),
  `RM-ANOVA p` = c(p_aov_treat, p_aov_time, p_aov_int),
  `LMM p`      = c(row_treat$p, row_time$p, row_int$p)
)
print(outcome)

# 6) Plot the data: spaghetti by bird + mean trend per treatment
# Helper for 95% CI around the mean
mean_ci <- function(y) {
  m <- mean(y)
  se <- sd(y) / sqrt(length(y))
  data.frame(y = m, ymin = m - 1.96 * se, ymax = m + 1.96 * se)
}

p <- ggplot(dat, aes(x = time, y = weight, color = treatment)) +
  # Thin lines per bird
  geom_line(aes(group = id), alpha = 0.25) +
  geom_point(alpha = 0.4, position = position_jitter(width = 0.05, height = 0)) +
  # Overlaid mean and 95% CI per treatment across time
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun.data = mean_ci, geom = "errorbar", width = 0.15, alpha = 0.7) +
  theme_bw() +
  labs(title = "Avian body weight over time by treatment",
       x = "Time (session)", y = "Body weight (g)")
print(p)
```


**Notes**

- The simulation creates a hierarchical, repeated-measures dataset (birds nested in treatment, measured over time) with random bird intercepts and treatment-specific time slopes to allow a Treatment × Time interaction.
- RM-ANOVA uses aov with Error(id/time), which assumes sphericity for within-subject effects; LMM relaxes this by modeling correlation via random effects.
- The outcome table reports p-values for Treatment main effect, Time main effect, and Treatment × Time interaction, side-by-side for RM-ANOVA and LMM.


### A case with differences



Below is a self-contained example that simulates a hierarchical repeated-measures dataset where RM-ANOVA and LMM/LME give different p-values, because:
- Within-subject correlation is AR(1) (violates sphericity).
- Residual variance changes over time (heteroscedasticity).
- Some missing observations (unbalanced).

The RM-ANOVA cannot model these features; mixed models can.

R code:

```{r}
# Packages
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(lme4)
  library(lmerTest)   # p-values for lmer
  library(nlme)       # correlation structures, heteroscedasticity
})

set.seed(42)

# 1) Simulate hierarchical repeated-measures data with AR(1), heteroscedasticity, and missingness
simulate_weight_complex <- function(
  group_names = c("Control","T1","T2","T3"),
  n_birds_per_group = c(15, 14, 16, 15),  # slightly unbalanced sample sizes
  n_time = 6,
  beta0 = 20,
  treat_intercepts = c(0, 1.2, 0.0, -1.0),      # treatment effects on intercept
  time_slope = 1.0,
  treat_slope_dev = c(0, 0.3, -0.2, -0.5),      # treatment-specific slope deviations
  sd_id_intercept = 2.0,                         # random intercept SD (bird)
  sd_id_slope = 0.8,                             # random slope SD (time)
  cor_id = 0.2,                                  # intercept-slope correlation
  rho_ar1 = 0.5,                                 # AR(1) residual corr within bird
  sd_time_vec = c(1.0, 1.2, 1.5, 1.9, 2.4, 3.0), # heteroscedastic residual SD over time
  miss_prop = 0.2,                                # random missingness of individual obs
  miss_last_time_prop = 0.3                       # extra dropout at the last time
) {
  stopifnot(length(group_names) == length(n_birds_per_group))
  stopifnot(length(sd_time_vec) == n_time)
  
  # Build id list by group
  ids <- unlist(mapply(function(g, n) paste0(g, "_", seq_len(n)), group_names, n_birds_per_group))
  id_df <- data.frame(
    id = ids,
    treatment = rep(group_names, times = n_birds_per_group),
    stringsAsFactors = FALSE
  )
  id_df$treatment <- factor(id_df$treatment, levels = group_names)
  
  # Random effects per bird: (intercept, slope) with correlation
  Sigma_b <- matrix(c(sd_id_intercept^2, cor_id*sd_id_intercept*sd_id_slope,
                      cor_id*sd_id_intercept*sd_id_slope, sd_id_slope^2), 2, 2)
  RE <- MASS::mvrnorm(n = nrow(id_df), mu = c(0, 0), Sigma = Sigma_b)
  colnames(RE) <- c("b0", "b1")
  id_df$b0 <- RE[, "b0"]
  id_df$b1 <- RE[, "b1"]
  
  # Expand to repeated measures
  df <- id_df %>%
    tidyr::uncount(n_time) %>%
    group_by(id) %>%
    mutate(time = row_number()) %>%
    ungroup() %>%
    mutate(time_c = time - mean(time),
           time_f = factor(time))
  
  # Fixed effects by treatment
  tr_idx <- match(df$treatment, group_names)
  tr_int <- treat_intercepts[tr_idx]
  tr_slp <- time_slope + treat_slope_dev[tr_idx]
  
  # Linear predictor with random intercept and random slope per bird (time as numeric)
  mu <- beta0 + tr_int + (tr_slp + df$b1) * df$time_c + df$b0
  
  # AR(1) residuals within each bird, then inflate SD by time-specific factor
  df <- df %>% arrange(id, time)
  df$eps <- NA_real_
  for (this_id in unique(df$id)) {
    idx <- which(df$id == this_id)
    Tn  <- length(idx)
    # Generate AR(1) errors with base SD = 1, then scale per time
    e <- numeric(Tn)
    z <- rnorm(Tn, 0, 1)
    e[1] <- z[1]
    if (Tn > 1) for (t in 2:Tn) e[t] <- rho_ar1 * e[t-1] + sqrt(1 - rho_ar1^2) * z[t]
    # Scale by heteroscedastic SD per time
    e <- e * sd_time_vec[df$time[idx]]
    df$eps[idx] <- e
  }
  
  df$weight <- mu + df$eps
  
  # Missingness: random missing + extra dropout at last time
  miss_rand <- runif(nrow(df)) < miss_prop
  miss_last <- (df$time == n_time) & (runif(nrow(df)) < miss_last_time_prop)
  df <- df[!(miss_rand | miss_last), ]
  
  df
}

dat <- simulate_weight_complex()

# 2) RM-ANOVA (requires complete cases across time for each subject)
dat_complete <- dat %>%
  group_by(id) %>%
  filter(n_distinct(time) == max(time)) %>%
  ungroup()

# Fit RM-ANOVA: between = treatment, within = time_f
aov_rm <- aov(weight ~ treatment * time_f + Error(id/time_f), data = dat_complete)
sum_aov <- summary(aov_rm)

# Helper to extract RM-ANOVA p-values
get_aov_p <- function(sum_list, stratum, term) {
  tab <- sum_list[[stratum]][[1]]
  rn <- rownames(tab)
  hit <- which(startsWith(rn, term))
  if (length(hit) == 0) return(NA_real_)
  as.numeric(tab[hit[1], "Pr(>F)"])
}
p_aov_treat <- get_aov_p(sum_aov, "Error: id", "treatment")
p_aov_time  <- get_aov_p(sum_aov, "Error: id:time_f", "time_f")
p_aov_int   <- get_aov_p(sum_aov, "Error: id:time_f", "treatment:time_f")

# 3) Mixed models

# 3a) LMM (lmer): random intercept + random slope for time (numeric)
# Use the subjects available in dat (including those with missing times)
options(contrasts = c("contr.sum", "contr.poly"))
m_lmer <- lmer(weight ~ treatment * time_c + (1 + time_c | id), data = dat)
anova_lmer <- anova(m_lmer, type = 3)  # Satterthwaite df

# 3b) LME (nlme): random intercept+ slope, AR(1) within-id, heteroscedastic residual SD by time
m_lme <- lme(
  fixed = weight ~ treatment * time_c,
  random = ~ time_c | id,
  correlation = corAR1(value = 0.5, form = ~ time | id),
  weights = varIdent(value = c("2"=1.2,"3"=1.5,"4"=1.9,"5"=2.4,"6"=3.0),form = ~ 1 | time_f),
  data = dat,
  method = "REML",
  control = lmeControl(msMaxIter = 200, msMaxEval = 400, niterEM = 50, tolerance = 1e-6, returnObject = TRUE)
)
m1 <- lme(weight ~ treatment * time_c, random = ~ time_c | id, data = dat)
m2 <- update(m1, correlation = corAR1(value = 0.5, form = ~ time | id))
m3 <- update(m2, weights = varPower(form = ~ time)) # fewer parameters, often easier
## m4 <- update(m2, varIdent(value = c("2"=1.2,"3"=1.5,"4"=1.9,"5"=2.4,"6"=3.0),form = ~ 1 | time_f)) # **does not work**
anova_lme <- anova(m_lme)

# Extract p-values (LMM from lmerTest, LME from nlme)
extract_lmer_p <- function(tab, term_label) as.numeric(tab[term_label, "Pr(>F)"])
p_lmer_treat <- extract_lmer_p(anova_lmer, "treatment")
p_lmer_time  <- extract_lmer_p(anova_lmer, "time_c")
p_lmer_int   <- extract_lmer_p(anova_lmer, "treatment:time_c")

extract_lme_p <- function(tab, i) as.numeric(tab$"p-value"[i])
# In nlme::anova, rows are: (Intercept), treatment, time_c, treatment:time_c
p_lme_treat <- extract_lme_p(anova_lme, 2)
p_lme_time  <- extract_lme_p(anova_lme, 3)
p_lme_int   <- extract_lme_p(anova_lme, 4)

# 4) Outcome table comparing p-values
outcome <- tibble::tibble(
  Effect = c("Treatment", "Time", "Treatment × Time"),
  `RM-ANOVA p` = c(p_aov_treat, p_aov_time, p_aov_int),
  `LMM (lmer) p` = c(p_lmer_treat, p_lmer_time, p_lmer_int),
  `LME (AR1+het) p` = c(p_lme_treat, p_lme_time, p_lme_int)
)
print(outcome)

# 5) Plot: spaghetti and mean trends
# Helper for 95% CI around the mean
mean_ci <- function(y) {
  m <- mean(y); se <- sd(y) / sqrt(length(y))
  data.frame(y = m, ymin = m - 1.96*se, ymax = m + 1.96*se)
}

p <- ggplot(dat, aes(x = time, y = weight, color = treatment)) +
  geom_line(aes(group = id), alpha = 0.25) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0.05, height = 0)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.1) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun.data = mean_ci, geom = "errorbar", width = 0.15, alpha = 0.7) +
  theme_bw() +
  labs(title = "Avian body weight over time by treatment",
  subtitle= "(AR(1), heteroscedastic, missing)",
       x = "Time", y = "Body weight (g)") +theme_bw(base_size=18)
print(p)
```


What to expect
- RM-ANOVA uses only subjects with complete time series and assumes sphericity and homoscedasticity. With AR(1) correlation, heteroscedasticity, and missing data, its p-values can be biased (often too liberal for within-subject effects).
- LMM (lmer) accounts for random intercepts and slopes; LME (nlme) additionally models AR(1) and time-varying residual SD. Their p-values typically differ from RM-ANOVA under these violations.
- You should see different p-values across methods, especially for Time and Treatment × Time.



# Cons of GLMM

Below is a minimal, self-contained R script that simulates a “small number of fields” case (n_fields = 4 with field-level treatment assignment), then compares treatment-effect p-values via:
- Satterthwaite (lmerTest, REML, Type III)
- Likelihood-ratio test (LRT; ML, nested model comparison)
- Parametric bootstrap (PB; pbkrtest)

This setup often yields noticeably different p-values because the true replication for the treatment effect is at the field level (2 vs 2), while the LRT can look overly optimistic when many within-field observations are present.


```{r}
# Packages
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(lme4)
  library(lmerTest)   # Satterthwaite df
  library(pbkrtest)   # Kenward-Roger and parametric bootstrap comparisons
})

set.seed(2025)

# 1) Simulate: 4 fields (clusters), treatment assigned at field level (2 Control, 2 Treated)
simulate_small_clusters <- function(
  n_fields = 4,
  fields_per_trt = c(Control = 2, Treated = 2),
  n_sessions = 10,
  n_reps_per_session = 10,  # within-field replication
  beta0 = 20,               # overall intercept (e.g., body weight)
  beta_treat = 1.0,         # treatment effect (Treated - Control) on response scale
  sd_field = 2.0,           # random field SD (between-field variability)
  sd_resid = 0.3            # residual SD (within-field variability)
) {
  stopifnot(sum(fields_per_trt) == n_fields)
  
  field <- factor(paste0("F", seq_len(n_fields)))
  treatment_by_field <- c(rep("Control", fields_per_trt["Control"]),
                          rep("Treated", fields_per_trt["Treated"]))
  treatment_by_field <- factor(treatment_by_field, levels = c("Control","Treated"))
  
  # Build design: sessions x reps within each field
  df <- do.call(rbind, lapply(seq_len(n_fields), function(i) {
    expand.grid(
      field = field[i],
      session = seq_len(n_sessions),
      rep = seq_len(n_reps_per_session),
      KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
    )
  })) %>%
    as.data.frame()
  
  df$treatment <- treatment_by_field[as.integer(factor(df$field, levels = field))]
  df$treatment <- factor(df$treatment, levels = c("Control","Treated"))
  
  # Random field effects
  b_field <- rnorm(n_fields, 0, sd_field); names(b_field) <- field
  
  # Linear predictor: y = beta0 + beta_treat * I(Treated) + b_field + e
  eta <- beta0 +
    (df$treatment == "Treated") * beta_treat +
    b_field[as.character(df$field)]
  
  y <- rnorm(nrow(df), mean = eta, sd = sd_resid)
  df$y <- y
  df
}

dat <- simulate_small_clusters()

# 2) Plot: raw points by field, with group means
ggplot(dat, aes(x = field, y = y, color = treatment)) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "black") +
  stat_summary(fun = mean, geom = "point", size = 2) +
  theme_bw() +
  labs(title = "Simulated data: 4 fields",
  subtitle ="many within-field observations",
       y = "Response", x = "Field") + theme_bw(base_size=18)

# 3) Fit LMMs
# Satterthwaite/KR generally use REML; LRT uses ML
m_full_REML <- lmer(y ~ treatment + (1 | field), data = dat, REML = TRUE)
m_null_REML <- lmer(y ~ 1 + (1 | field), data = dat, REML = TRUE)

m_full_ML <- lmer(y ~ treatment + (1 | field), data = dat, REML = FALSE)
m_null_ML <- lmer(y ~ 1 + (1 | field), data = dat, REML = FALSE)

# 4) Satterthwaite p-value (fixed effect of treatment)
# Use Type III with sum-to-zero contrasts for clarity (optional)
options(contrasts = c("contr.sum","contr.poly"))
anova_satt <- anova(m_full_REML, type = 3)
p_satt <- anova_satt["treatment", "Pr(>F)"]
df_satt <- anova_satt["treatment", "DenDF"]

# 5) LRT p-value (ML fit; full vs null)
lrt <- anova(m_null_ML, m_full_ML, test = "Chisq")
p_lrt <- lrt$`Pr(>Chisq)`[2]
chi <- lrt$Chisq[2]
df_lrt <- lrt$Df[2]

# 6) Parametric bootstrap p-value (pbkrtest)
# This resamples under the null and compares the full-null statistic; increase nsim for stability.
set.seed(2026)
pb <- PBmodcomp(m_full_REML, m_null_REML, nsim = 1000, seed = 2026)
# PBmodcomp returns different tests; we take the "PBtest" line
p_pb <- pb$test["PBtest", "p.value"]

# 7) Collect results
outcome <- data.frame(
  Method = c("Satterthwaite (REML, Type III)", "LRT (ML, full vs null)", "Parametric bootstrap (PB)"),
  Statistic = c(paste0("DenDF≈", round(df_satt, 2)),
                paste0("ChiSq=", round(chi, 2), ", df=", df_lrt),
                paste0("nsim=", attr(pb, "nsim"))),
  `p-value` = signif(c(p_satt, p_lrt, p_pb), 4),
  check.names = FALSE
)
print(outcome)

# 8) Optional: show field-level means (to emphasize true replication)
field_means <- dat %>%
  group_by(field, treatment) %>%
  summarize(mean_y = mean(y), .groups = "drop")
print(field_means)
```


**Notes on why big differences can often be seen here**

- Treatment is assigned at the field level (2 control vs 2 treated), so the effective replication for the treatment effect is the number of fields, not the number of within-field measurements.
- Satterthwaite (and Kenward–Roger) base the denominator df on cluster-level information; with 4 fields, DenDF is tiny (often around 2), yielding larger p-values.
- The ML LRT can look much more optimistic because it leverages the large number of within-field observations in the likelihood. In small-cluster settings, that can understate uncertainty about a cluster-level treatment effect.
- Parametric bootstrap (PB) tends to align more with small-sample cluster-level uncertainty, often closer to Satterthwaite/KR than to the LRT, especially with very few clusters.

If the differences are not large on a particular run, tweak the simulation to accentuate the effect:

- Increase within-field replication (n_sessions, n_reps_per_session) and reduce residual SD (sd_resid) to make LRT smaller.
- Increase between-field SD (sd_field) so cluster differences dominate, keeping Satterthwaite/PB more conservative.
- Keep only 4 fields with 2 per treatment to maintain minimal cluster replication.


## Separation Problem


I have a dataset to demonstrate (quasi-)complete separation. In the 48 h data, Control, 5, 10, and 20 mg/L have zero immobility in every replicate, while 80 mg/L has high immobility and 40 mg/L has some. A logistic GLM with Treatment as a factor will hit separation: some treatment levels have all 0s, others have mostly 1s. You’ll typically see “fitted probabilities 0 or 1 occurred,” huge SEs, and unstable p-values.

Here’s a concise workflow to show it, then fix it with bias-reduced logistic regression.

R code

```{r}
# Using your 'dat' object (48 h table with columns: Treatment, Dose, Immobile, Total)

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(brglm2)    # bias-reduced logistic + separation detection
  library(broom)
})

# 1) Show the pattern that causes separation
by_trt <- dat %>%
  group_by(Treatment) %>%
  summarize(total_immob = sum(Immobile), total_n = sum(Total),
            prop_immob = total_immob / total_n, .groups = "drop")
print(by_trt)

# Visual: proportion immobile by treatment (points with 95% binomial CI)
ggplot(by_trt, aes(x = factor(Treatment,levels=c("Control","5",10,20,40,80)), y = prop_immob)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = prop_immob - 1.96*sqrt(prop_immob*(1-prop_immob)/total_n),
                    ymax = prop_immob + 1.96*sqrt(prop_immob*(1-prop_immob)/total_n)),
                width = 0.2) +
  theme_bw() +
  labs(x = "Treatment", y = "Proportion immobile", title = "Quantal responses by treatment (separation visible)")

# 2) Fit logistic GLM with Treatment as factor (expect separation warnings)
m_glm_trt <- glm(cbind(Immobile, Total - Immobile) ~ Treatment,
                 data = dat, family = binomial)
summary(m_glm_trt)
# You’ll likely see "fitted probabilities numerically 0 or 1 occurred" and very large SEs.

# 3) Detect separation formally

library(detectseparation)
sep <-    glm(cbind(Immobile, Total - Immobile) ~ Treatment,
                                 data = dat, family = binomial("logit"), method = "detect_separation")
print(sep)
# The report indicates which coefficients diverge (complete or quasi-complete separation).

# 4) Bias-reduced (penalized) logistic to obtain finite, stable estimates
m_br_trt <- glm(cbind(Immobile, Total - Immobile) ~ Treatment,
                          data = dat, family = binomial("logit"),
                          method ="brglmFit",type = "AS_median")
summary(m_br_trt)

# 5) Trend test alternative (numeric dose) can still struggle if there’s no overlap.
# Use log10 dose; add a tiny constant to allow log(0) if needed.
dat <- dat %>% mutate(logdose = log10(pmax(Dose, 1e-6)))
m_glm_trend <- glm(cbind(Immobile, Total - Immobile) ~ logdose,
                   data = dat, family = binomial)
summary(m_glm_trend)

# Bias-reduced trend fit
m_br_trend <- glm(cbind(Immobile, Total - Immobile) ~ logdose,
                            data = dat, family = binomial("logit"),
                            method ="brglmFit", type = "AS_median")
summary(m_br_trend)

# 6) Plot fitted curves (classical GLM vs bias-reduced) over raw proportions
# Aggregate proportions per dose for plotting
agg <- dat %>% group_by(Dose) %>%
  summarize(prop = sum(Immobile)/sum(Total), n = sum(Total), .groups = "drop")
newd <- data.frame(Dose = sort(unique(dat$Dose)))
newd$logdose <- log10(pmax(newd$Dose, 1e-6))
newd$pred_glm <- plogis(predict(m_glm_trend, newdata = newd, type = "link"))
newd$pred_br  <- plogis(predict(m_br_trend,  newdata = newd, type = "link"))

ggplot(agg, aes(Dose, prop)) +
  geom_point(size = 2) +
  geom_line(data = newd, aes(Dose, pred_glm, color = "GLM (MLE)")) +
  geom_line(data = newd, aes(Dose, pred_br,  color = "Bias-reduced (brglm2)"), linetype = "dashed") +
  scale_color_manual(values = c("GLM (MLE)" = "#1f77b4", "Bias-reduced (brglm2)" = "#d62728")) +
  theme_bw() +
  labs(y = "Proportion immobile", color = "Fit",
       title = "Dose–response logistic fits: MLE vs bias-reduced")
models <- list(m_glm_trt,m_br_trt)
names(models) <- c("GLM","Reduced-bias")
modelsummary::modelsummary(models,estimate = "{estimate} ({std.error}){stars}",statistic =NULL)
```


- Why separation occurs here
  - Several treatments have 0 immobility in every replicate, while high dose has consistent immobility; the design perfectly (or quasi-perfectly) separates outcomes by treatment/dose.
- Symptoms in GLM
  - Warnings about fitted probabilities 0 or 1, huge standard errors, unstable or infinite estimates, and misleading p-values.
- Remedies
  - Bias-reduced/Firth logistic (brglm2::brglm, logistf) produces finite, stable estimates and CIs.
  - Consider modeling dose as a numeric covariate (log-dose), but if overlap is absent, bias-reduction is still recommended.
  - For extra-binomial variation, a beta-binomial or a binomial GLMM with an observation-level random effect can address overdispersion—but note: GLMMs do not “fix” separation by themselves; separation can persist in mixed models.

**GLMM note**
- If you try a binomial GLMM like glmer(cbind(Immobile, Total - Immobile) ~ logdose + (1|Replicate), family=binomial, data=dat), you may still get convergence warnings or boundary estimates when separation is near-perfect. Penalized methods (e.g., brglm2) are designed for this issue.



# Links

Here are concise, reliable links you can add to a references slide for GLMM/GAMM work in ecotoxicology:

**Core papers and books**
- Bates, Mächler, Bolker, Walker (2015). Fitting linear mixed-effects models using lme4 (JSS): https://www.jstatsoft.org/article/view/v067i01
- Zuur et al. (2009). Mixed Effects Models and Extensions in Ecology with R (Springer): https://link.springer.com/book/10.1007/978-0-387-87458-6
- Bolker (2008). Ecological Models and Data in R (Princeton): https://press.princeton.edu/books/paperback/9780691125229/ecological-models-and-data-in-r
- Harrison et al. (2018). A brief introduction to mixed effects modelling in ecology (MEE): https://doi.org/10.1111/2041-210X.12570
- Wood (2017). Generalized Additive Models: An Introduction with R (CRC): https://www.routledge.com/Generalized-Additive-Models/WOOD/p/book/9781498728331

**R packages and vignettes**

- lme4 (LMM/GLMM): https://cran.r-project.org/package=lme4
- glmmTMB (GLMM incl. NB, zero-inflation): https://cran.r-project.org/package=glmmTMB
  - Vignette: https://cran.r-project.org/web/packages/glmmTMB/vignettes/glmmTMB.pdf
- mgcv (GAM/GAMM): https://cran.r-project.org/package=mgcv
  - Intro vignette: https://cran.r-project.org/web/packages/mgcv/vignettes/mgcv.pdf
- DHARMa (diagnostics for GLM/GLMM): https://cran.r-project.org/package=DHARMa
  - Vignette: https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html
- pbkrtest (Kenward–Roger, parametric bootstrap): https://cran.r-project.org/package=pbkrtest
- emmeans (marginal means/contrasts): https://cran.r-project.org/package=emmeans
- MuMIn (AICc, model selection/averaging): https://cran.r-project.org/package=MuMIn
- simr (power analysis for LMM/GLMM): https://cran.r-project.org/package=simr
- brglm2 (bias-reduced GLMs; separation diagnostics): https://cran.r-project.org/package=brglm2

**Practical guides and FAQs**

- Ben Bolker’s GLMM FAQ and mixed-model notes: https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html
- GLMM troubleshooting/overdispersion (glmmTMB vignette): https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.pdf
- Performance package (model checks, collinearity, residuals): https://cran.r-project.org/package=performance


These cover fitting, diagnostics, inference, model selection, and power—everything you need to support GLMM/GAMM analyses in ecotoxicology.