---
title: "Pseudo $R^2$"
description: "A discussion on confidence intervals"
author: "Zhenglei Gao"
date: "11/20/2024"
categories:
  - code
  - stats
  - R
editor_options: 
  chunk_output_type: console
---

When discussing with Johannes, I mentioned that GoF is making sense when comparing with other models, most of the time with Null or full models.

## General considerations


When used in nonlinear regression, the pseudo $R^2$ cannot be interpreted as in linear regression. 

## $R^2$ can be misleading


```{r}
x <- rnorm(100)
y <- 3*x+rnorm(100)
y1 <- y-0.1
y2 <- y- 1000+rnorm(100)
summary(lm(y~x))  ## R^2 = 0.8855
summary(lm(y~y1)) ## R2 = 1 ## Is this a good fit, no. It is a biased fit

summary(lm(y~y2)) ## R2 = 1 ## Is this a good fit, no. It is a biased fit
plot(y,y1,ylim=c(-1000,10))
points(y,y2,col="red")


x <- seq(1,100,1)
length(x)
y <- 3*x+rnorm(length(x),0,5)

summary(lm(y~x))

y1 <- y*0.9+rnorm(length(x),0,0.11) ## systematic error!
plot(x,y)
points(x,y1,col="red")
plot(y,y1)
summary(lm(y~y1))
summary(lm(y~x))
ypred <- predict(lm(y~x))
summary(lm(y~ypred))
```




## Try with some examples

From the pscl package: 

- llh: The log-likelihood from the fitted model
- llhNull: The log-likelihood from the intercept-only restricted model
- G2: Minus two times the difference in the log-likelihoods
- McFadden: McFadden's pseudo r-squared
- r2ML: Maximum likelihood pseudo r-squared
- r2CU: Cragg and Uhler's pseudo r-squared

### Example with random data

```{r}
# Load necessary libraries
library(MASS)
## install.packages("pscl")
library(pscl)  # For the pseudo R-squared function

# Simulate some data
set.seed(123)
n <- 100
x <- rnorm(n)
y <- rbinom(n, 1, prob = 0.5)  # Random binary outcome that is not related to x

# Fit a logistic regression model
model <- glm(y ~ x, family = binomial)

# Calculate pseudo R-squared
pseudo_r2 <- pR2(model)
print(pseudo_r2)
summary(lm(y~predict(model)))
# Check the model summary
summary(model)

# Visualize the fit
plot(x, y, main = "Logistic Regression Fit", xlab = "x", ylab = "y")
points(x, fitted(model), col = "red", pch = 19)
```

### Example where R2 works 

```{r}
# Simulate data with outliers
# Load necessary libraries
library(MASS)
library(pscl)  # For the pseudo R-squared function

# Simulate some data
set.seed(456)
n <- 100
x <- rnorm(n)
x0 <- x
# Create a binary outcome based on a logistic function
y <- rbinom(n, 1, prob = plogis(x))
y0 <- y
# Introduce influential points by adding extreme values to the predictor
r1 <- order(x)
ind <- r1[91:100]
ind <- r1[1:10]
x[ind] <- rep(3, 10)  # Make these points influential by shifting x

# Recalculate the binary response based on the new x values
y[ind] <- rbinom(10, 1, prob = plogis(x[ind]))  # Ensure y is binary

# Fit a logistic regression model
model_with_influential_points <- glm(y ~ x0, family = binomial)
summary(lm(y~predict(model_with_influential_points)))
model_without_influential_points <- glm(y ~ x, family = binomial)
summary(lm(y0~predict(model_without_influential_points)))
# Calculate pseudo R-squared
pseudo_r2_outliers <- pR2(model_with_influential_points)
print(pseudo_r2_outliers)
rbind(pseudo_r2_outliers,pR2(model_without_influential_points))
# Check the model summary
summary(model_with_influential_points)
# Predict probabilities
predicted_probabilities <- predict(model_with_influential_points, type = "response")

# Classify predictions (using 0.5 as the cutoff)
predicted_classes <- ifelse(predicted_probabilities > 0.5, 1, 0)

# Calculate classification error
classification_error <- mean(predicted_classes != y)
print(paste("Classification Error:", round(classification_error, 4)))


# Predict probabilities
predicted_probabilities <- predict(model_without_influential_points, type = "response")

# Classify predictions (using 0.5 as the cutoff)
predicted_classes <- ifelse(predicted_probabilities > 0.5, 1, 0)

# Calculate classification error
classification_error <- mean(predicted_classes != y0)
print(paste("Classification Error:", round(classification_error, 4)))
# Visualize the fit
plot(x0, y, main = "Logistic Regression Fit with Influential Points", xlab = "x", ylab = "y", ylim = c(-0.1, 1.1))
points(x0, fitted(model_with_influential_points), col = "green", pch = 3)
points(x0[ind], y0[ind]-0.1, col = "red", pch = 3)
points(x0[ind], y[ind]-0.05, col = "darkgreen", pch = 4)
points(x0, fitted(model_without_influential_points), col = "red", pch = 19)
```


## A case where R^2 can be 0 to 1 depending on the data.

depending on x is between (-3,3), (-3,6), (-1,3)!



```{r}
# Create data that does not fit a logistic model well
set.seed(789)
n <- 100
x <- seq(-3, 6, length.out = n)
y <- ifelse(x^2 > 1, 1, 0)  # Nonlinear relationship
y1 <- ifelse(x > 0, 1, 0)  # Perfect fit
# Fit a logistic regression model
model_nonlinear <- glm(y ~ x, family = binomial)

# Calculate pseudo R-squared
pseudo_r2_nonlinear <- pR2(model_nonlinear)
print(pseudo_r2_nonlinear)
xa <- rep(1,length(y))
summary(lm(y~predict(model_nonlinear)))
# Check the model summary
summary(model_nonlinear)

# Visualize the fit
plot(x, y, main = "Logistic Regression Fit for Nonlinear Data", xlab = "x", ylab = "y")
points(x, fitted(model_nonlinear), col = "red", pch = 19)
```


## Some Additional Notes

- https://rcompanion.org/rcompanion/b_04.html

## G-test = Likelihood ratio test?

```{r}
observed = c(1752, 1895)    # observed frequencies
expected = c(0.5, 0.5)      # expected proportions

library(DescTools)

GTest(x=observed,
      p=expected,
      correct="none") 
degrees = 1                       # degrees of freedom



observed      = c(1752, 1895)     # observed frequencies
expected.prop = c(0.5, 0.5)       # expected proportions
expected.count = sum(observed)*expected.prop

G = 2 * sum(observed * log(observed / expected.count))

pchisq(G,
       df=degrees,
       lower.tail=FALSE)
```

### $R^2$ can only increase when adding more variables

```{r}

```



## An error in reporting R2.

The null mean model is different in those two types of model specifications, even though they are the same model.

::: callout-note
r.squared:  R^2, the ‘fraction of variance explained by the model’,
              R^2 = 1 - Sum(R[i]^2) / Sum((y[i]- y*)^2),
            where y* is the mean of y[i] if there is an intercept and zero otherwise.
::: 

```{r}
library("sigr")
library("broom")
set.seed(23255)

d <- data.frame(y= runif(100),
                x= ifelse(runif(100)>=0.5, 'a', 'b'),
                stringsAsFactors = FALSE)
m1 <- lm(y~x, data=d)
t(broom::glance(m1))
d$pred1 <- predict(m1, newdata = d)
cat(render(sigr::wrapFTest(d, 'pred1', 'y'),
           format='markdown'))
m2 <- lm(y~0+x, data=d)
t(broom::glance(m2))
```

- https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0183250

