---
title: "R Pipeline operators"
description: "magrittr and own definitions"
author: "Zhenglei Gao"
date: "06/12/2025"
categories:
  - R
editor_options: 
  chunk_output_type: console
---


::: callout-warning
To be written.
:::


## What I often use

### `%||%`

Quoted from R base control Help documentation: 
> The null coalescing operator `%||%` is a simple 1-line function: x %||% y is an idiomatic way to call

    if (is.null(x)) y else x
                             # or equivalently, of course,
    if(!is.null(x)) x else y 
Inspired by Ruby, it was first proposed by Hadley Wickham.

```{r}
res <- {}
res %||% "alternative result"
```



```{r}
library(magrittr)
```

## Forward-pipe operator

R now has `|>` in base. 

```{r}
rnorm(5000) %>%
  abs %>% `*` (100)  %>%
  matrix(ncol=10)  %>%
  rowMeans %>% round %>% 
  `%%`(9) %>% hist
```

## Tee operator (left) `%T>%`

```{r}
rnorm(5000) %>%
  abs %>% `*` (100)  %>%
  matrix(ncol=10)  %>%
  rowMeans %>% round %>% 
  `%%`(9) %T>% hist %>% sum
```

## Exposition pipe-operator

```{r}

```

## Compound assignment pipe-operator

