---
title: "Risk Report and Metric"
description: "Reproduce "
author: "Zhenglei Gao"
date: "12/06/2024"
categories:
  - code
  - Stats
  - R
editor_options: 
  chunk_output_type: console
---

Since I started working in an internal project that utilizes a validated R envrionment provided by a 3rd vendor company, I started 

## Testing the Idea

1. Install the package
```{r eval=FALSE}
pak::pak("pharmaR/riskreports")
```

2. Start generating a report

The **riskreports** package is developed by the R validation hub repo working group.  

```{r eval=FALSE}
library(riskreports)
options("riskreports_output_dir" = ".")
pr <- package_report(
  package_name = "dplyr",
  package_version = "1.1.4",
  params = list(
    assessment_path = system.file("assessments/dplyr.rds", package = "riskreports")),
  quiet = TRUE # To silence quarto output for readability
)
```


## Generate a summary report

I would prefer having also a summarize report in the end for the QA to check. I don't think the QA would check every single validation report for the handreds of packages included in the image. 

```{r}
tmp_packages_file <- tempfile("packages")
repo_packages_url <- "https://raw.githubusercontent.com/pharmaR/pharmapkgs/refs/heads/main/inst/repos/src/contrib/PACKAGES"
download.file(repo_packages_url, tmp_packages_file)
tmp <- read.dcf(tmp_packages_file)
colnames(tmp)
```

```{r}
library(dplyr)
knitr::kable(t(head(tmp))) %>% kableExtra::kable_classic(.) 
```



## Testing pharmapkgs

```{r eval=FALSE}
repo <- "C:/Projects/Rlib/pharmapkgs/repos/src/contrib/"
packages <- get_packages(
      base_url = system.file("repos", package = "pharmapkgs", mustWork = TRUE)
    )

available.packages(repos = repo, fields = risk_fields(repo)) |>
  tibble::as_tibble()


```
