---
title: "Set up GitHub Codespace"
description: "Generic steps to make the codespace my style"
author: "Zhenglei Gao"
date: "09/11/2025"
categories:
  - code
  - R
editor_options: 
  chunk_output_type: console
---

## Basics for R

- Install R and Quarto

```bash
sudo apt-get update && sudo apt-get install -y r-base
wget https://quarto.org/download/latest/quarto-linux-amd64.deb
sudo apt-get install -y ./quarto-linux-amd64.deb
rm quarto-linux-amd64.deb
```

- Install Extension: REditorSupport
    - install the requred packages: `sudo Rscript -e "install.packages('languageserver', repos='https://cloud.r-project.org/')"`
    `sudo Rscript -e "install.packages('httpgd', repos='https://cloud.r-project.org/')"`
    `sudo Rscript -e "install.packages('devtools', repos='https://cloud.r-project.org/')"`


- Install often used R packages


Note for certain packages, some additional system packages are required. From example, for PMCMRplus package, libmpfr need to be installed.

```bash
sudo apt update
sudo apt install libmpfr-dev
```


```{r}
#| eval: false
## Example: dependencies for drcHelper
install.packages("Rmpfr")
install.packages(c(
  "drc", "multcomp", "PMCMRplus", "dplyr", "tidyselect", "lme4", "bmd",
  "ggplot2", "scales", "metafor", "rstatix", "DescTools", "isotone"
))
install.packages("remotes") # if not already installed

remotes::install_github("DoseResponse/bmd@a3bd9c6113e35d43e0990614469dabb10cd409bd")
remotes::install_github("DoseResponse/drc")
remotes::install_github("DoseResponse/drcData")
```

For rendering the pkgdown website, additional packages are neede

```{r}
#| eval: false

install.packages(c(
  "GLMcat", "kableExtra", "emmeans", "outliers", "drda", "pander", "ggnewscale", "patchwork", "ggprism", "plyr"
))
```


- Install radian

```bash
## Ubuntu 23.04 or above, my current code space is 4.24
lsb_release -a
sudo apt update
sudo apt install pipx
pipx ensurepath
```

```bash
# install released version
pipx install radian
# or the dev version
pipx install git+https://github.com/randy3k/radian
# to run radian
radian
```


## Get around CRAN checks

## Github pages publishing workflow