---
title: "Set up GitHub Codespace"
description: "Generic steps to make the codespace my style"
author: "Zhenglei Gao"
date: "09/11/2025"
categories:
  - code
  - R
editor_options: 
  chunk_output_type: console
---

## Easiest Way

The easiest way to set up a GitHub Codespace with your preferred configuration is to use a `.devcontainer` configuration. This approach lets you define your development environment as code:

1. Create a `.devcontainer` folder in your repository
2. Add a `devcontainer.json` file with your configuration

For example, the [eddelbuettel/r2u repository](https://github.com/eddelbuettel/r2u) provides a ready-to-use configuration when launching a Codespace from the repository. You can find detailed instructions in their [Codespaces documentation](https://eddelbuettel.github.io/r2u/vignettes/Codespaces/).


--- 

Here's a sample configuration for R and Quarto manually set by my self:


## Basics for R

- Install R and Quarto

```bash
sudo apt-get update && sudo apt-get install -y r-base
```
```{bash}
#| eval: false
wget https://quarto.org/download/latest/quarto-linux-amd64.deb
sudo apt-get install -y ./quarto-linux-amd64.deb
rm quarto-linux-amd64.deb
```

- Install Pandoc

```{bash}
#| eval: false
sudo apt-get update && sudo apt-get install -y pandoc
```


- Install Extension: REditorSupport
    - install the requred packages: `sudo Rscript -e "install.packages('languageserver', repos='https://cloud.r-project.org/')"`
    `sudo Rscript -e "install.packages('httpgd', repos='https://cloud.r-project.org/')"`
   

- Install often used R packages

**Before installing the devtools R package on a fresh Ubuntu system, you must:**

Install system libraries required by devtools and its dependencies:
- libharfbuzz-dev
- libfribidi-dev
- libfreetype-dev
- libpng-dev
- libtiff-dev
- libwebp-dev
- libjbig-dev
- libdeflate-dev
- liblerc-dev
- libsharpyuv-dev

Use sudo when installing R packages to avoid permission issues.
These steps ensure all necessary C/C++ libraries are available for compiling devtools and its dependencies

Note for certain packages, some additional system packages are required. From example, for PMCMRplus package, libmpfr need to be installed.

```bash
sudo apt update
sudo apt install libmpfr-dev
```

 `sudo Rscript -e "install.packages('devtools', repos='https://cloud.r-project.org/')"`


```{r eval=FALSE}
#| eval: false
## Example: dependencies for drcHelper
install.packages("Rmpfr")
install.packages(c(
   "multcomp", "PMCMRplus", "dplyr", "tidyselect", "lme4", "bmd",
  "ggplot2", "scales", "metafor", "rstatix", "DescTools", "isotone"
))
install.packages("remotes") # if not already installed

remotes::install_github("DoseResponse/bmd@a3bd9c6113e35d43e0990614469dabb10cd409bd")
remotes::install_github("DoseResponse/drc")
remotes::install_github("DoseResponse/drcData")
```

To clean up lock files and install a package: `rm -f /home/codespace/R/x86_64-pc-linux-gnu-library/4.3/00LOCK-ragg && Rscript -e "install.packages('ragg', repos='https://cloud.r-project.org/')"`

For rendering the pkgdown website, additional packages are neede

```{r}
#| eval: false

install.packages(c(
  "GLMcat", "kableExtra", "emmeans", "outliers", "drda", "pander", "ggnewscale", "patchwork", "ggprism", "plyr","MCPMpd"
))
```

## Setting Local R library

```{r eval=FALSE}
dir.create(Sys.getenv("R_LIBS_USER"), recursive = TRUE, showWarnings = FALSE)
.libPaths(Sys.getenv("R_LIBS_USER"))
devtools::install()
```


## Install radian

```bash
## Ubuntu 23.04 or above, my current code space is 4.24
lsb_release -a
sudo apt update
sudo apt install pipx
pipx ensurepath
```

```bash
# install released version
pipx install radian
# or the dev version
pipx install git+https://github.com/randy3k/radian
# to run radian
radian
```

To find out the path of radian: `which radian`.

To associate R code execution in VS Code with radian instead of the default R terminal, follow these steps:

1. Install the VS Code R extension (if not already installed).
2. Set radian as the R terminal in VS Code settings:
  - Open VS Code settings (Ctrl+,).
  - Search for r.rterm.linux.
  - Set its value to the path of radian


## Install R 4.5 or higher

Here are step-by-step instructions to upgrade R to version 4.5 or higher on Ubuntu 24.04:

Remove the old R version (optional):

```{bash}
#| eval: false
sudo apt remove --purge r-base r-base-core r-recommended r-base-dev
```

Install required dependencies:


```{bash}
#| eval: false
sudo apt update
sudo apt install -y --no-install-recommends software-properties-common dirmngr gnupg apt-transport-https ca-certificates
```

Add the CRAN GPG key:

```{bash}
#| eval: false
sudo apt-key adv --keyserver keyserver.ubuntu.com --recv-keys 'E298A3A825C0D65DFD57CBB651716619E084DAB9'
```


Add the CRAN repository for R (for Ubuntu 24.04 "Noble"):

```{bash}
#| eval: false
echo "deb https://cloud.r-project.org/bin/linux/ubuntu noble-cran40/" | sudo tee /etc/apt/sources.list.d/cran.list
```

Update package list:

```{bash}
#| eval: false
sudo apt update
```

Install the latest R version:

```{bash}
#| eval: false
sudo apt install -y r-base
```

Verify the R version:
```{bash}
#| eval: false
R --version
```

This should install R 4.5.x or newer. 

## Get around CRAN checks

## Github pages publishing workflow


## Prompts instruct the GitHub Agent

```
- When there are syntax errors that cannot be easily fixed, the issue seems often incomplete or fragmented code. Write clean versions instead, also, try to delete the version non-clean afterwards with the confirmation from me. 
- 
```